---
title: "Simulation Benchmark Data Generation"
author: "Tanmay Tanna <ttanna@ethz.ch>"
output:
  rmarkdown::html_notebook:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 180
)

ensure_package <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to load required package: ", pkg)
  }
}

for (pkg in c("dplyr", "ggplot2", "knitr", "readr", "scales", "tibble", "tidyr")) {
  ensure_package(pkg)
}
```

This notebook reproduces the shareable simulation data generation workflow described in the manuscript sections **Simulation benchmark** and **Supplementary Methods S6**. It creates the clean ground-truth matrix, adds metabolite-specific batch and drift effects, and writes the benchmark inputs used for downstream method comparisons.

## 1. Generate the manuscript-aligned simulation benchmark

```{r run-simulation}
source(file.path("..", "scripts", "simulate_metabolomics_benchmark.R"))

simulation <- simulate_metabolomics_benchmark(
  output_dir = file.path("..", "data", "simulated"),
  verbose = FALSE
)

meta <- simulation$metadata
design_table <- simulation$design_table
summary_values <- simulation$summary
```

## 2. Cohort layout

```{r cohort-summary}
cohort_summary <- tibble::tibble(
  Metric = c(
    "Metabolites",
    "Total injections",
    "Plates",
    "Controls",
    "Study samples",
    "QC interval per plate"
  ),
  Value = c(
    summary_values$n_metabolites,
    summary_values$n_injections,
    summary_values$n_plates,
    summary_values$controls,
    summary_values$study_samples,
    summary_values$qc_interval
  )
)

knitr::kable(cohort_summary, align = c("l", "r"))
```

```{r layout-plot}
layout_plot <- ggplot2::ggplot(
  meta,
  ggplot2::aes(
    x = order_in_plate,
    y = factor(plate, levels = rev(unique(plate))),
    fill = sample_type
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.25) +
  ggplot2::scale_fill_manual(values = c(control = "#1f77b4", sample = "#d9d9d9")) +
  ggplot2::labs(
    title = "Simulation acquisition layout",
    subtitle = "Five plates of 100 injections with pooled controls inserted every 10 acquisitions.",
    x = "Order within plate",
    y = "Plate",
    fill = "Injection type"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(panel.grid = ggplot2::element_blank())

layout_plot
```

## 3. Injected technical effects

```{r technical-summary}
effect_summary <- tibble::tibble(
  Metric = c(
    "Metabolites with any batch effect",
    "Strong-shift subset among batch-affected metabolites",
    "Metabolites with drift on at least one plate",
    "Mean drift-active metabolite-plate pairs"
  ),
  Value = c(
    sprintf("%.1f%%", summary_values$batch_affected_pct),
    sprintf("%.1f%%", summary_values$strong_batch_subset_pct),
    sprintf("%.1f%%", 100 * mean(design_table$drift_effect_applied_any_plate)),
    sprintf("%.1f%%", summary_values$drift_active_pair_pct)
  )
)

knitr::kable(effect_summary, align = c("l", "r"))
```

```{r drift-type-summary}
drift_type_counts <- design_table |>
  dplyr::summarise(
    linear = sum(drift_type_linear),
    nonlinear = sum(drift_type_nonlinear),
    ar1 = sum(drift_type_ar1),
    mixed = sum(drift_type_mixed)
  ) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "drift_type", values_to = "count")

drift_type_plot <- ggplot2::ggplot(
  drift_type_counts,
  ggplot2::aes(x = drift_type, y = count, fill = drift_type)
) +
  ggplot2::geom_col(width = 0.7, color = "#1f1f1f", linewidth = 0.25) +
  ggplot2::scale_fill_manual(
    values = c(
      linear = "#4e79a7",
      nonlinear = "#59a14f",
      ar1 = "#e15759",
      mixed = "#f28e2b"
    ),
    guide = "none"
  ) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(
    title = "Assigned drift morphologies across active metabolite-plate pairs",
    x = "Drift type",
    y = "Count"
  ) +
  ggplot2::theme_minimal(base_size = 12)

drift_type_plot
```

## 4. Saved benchmark files

```{r saved-files}
saved_files <- tibble::tibble(
  File = basename(simulation$output_paths),
  Path = unname(simulation$output_paths)
)

knitr::kable(saved_files, align = c("l", "l"))
```

The written files provide the shareable synthetic benchmark used in the manuscript: a clean reference matrix, an artefact-bearing matrix, sample metadata, and a feature-level design audit table.
