---
title: "Public MTBLS79 Preprocessing and Imputation"
author: "Tanmay Tanna <ttanna@ethz.ch>"
output:
  rmarkdown::html_notebook:
    toc: true
    number_sections: true
params:
  dataset_path: "../data/public/raw/Dataset07__SFPM.xlsx"
  download_if_missing: false
  direct_download_url: "https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS79/FILES/Dataset07__SFPM.xlsx"
  record_url: "https://www.ebi.ac.uk/metabolights/MTBLS79"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 180
)

ensure_package <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to load required package: ", pkg)
  }
}

for (pkg in c("dplyr", "ggplot2", "knitr", "missForest", "openxlsx", "scales", "tidyr")) {
  ensure_package(pkg)
}
```

This notebook reproduces the shareable public-data preprocessing described in **Public multi-batch measurement benchmark** and **Supplementary Methods S7**. It expects the public MetaboLights file `Dataset07__SFPM.xlsx`, applies the manuscript's missing-value handling procedure, and writes the processed CSV files used for downstream analysis.

Users should normally download the spreadsheet from the MTBLS79 record at [MetaboLights](https://www.ebi.ac.uk/metabolights/MTBLS79) and place it at `data/public/raw/Dataset07__SFPM.xlsx`.

## 1. Resolve the input dataset

```{r resolve-input}
dataset_path <- params$dataset_path

if (!file.exists(dataset_path) && isTRUE(params$download_if_missing)) {
  dir.create(dirname(dataset_path), showWarnings = FALSE, recursive = TRUE)
  utils::download.file(
    url = params$direct_download_url,
    destfile = dataset_path,
    mode = "wb",
    method = "libcurl",
    quiet = TRUE
  )
}

if (!file.exists(dataset_path)) {
  stop(
    "Dataset07__SFPM.xlsx was not found at ", dataset_path, ".\n",
    "Download it from ", params$record_url, " and rerun the notebook."
  )
}
```

## 2. Run the manuscript preprocessing workflow

```{r run-preprocessing}
source(file.path("..", "scripts", "preprocess_mtbls79_public_data.R"))

public_data <- preprocess_mtbls79_public_data(
  file_path = dataset_path,
  output_dir = file.path("..", "data", "public", "processed"),
  seed = 42L,
  reuse_existing = TRUE,
  verbose = FALSE
)

meta <- public_data$raw_metadata
raw_data <- public_data$raw_data
imputed_data <- public_data$imputed_data
feature_summary <- public_data$feature_summary
```

## 3. Dataset summary

```{r feature-summary}
summary_table <- tibble::tibble(
  Metric = c(
    "Samples",
    "Raw features",
    "Retained after 80% rule",
    "Removed before imputation"
  ),
  Value = c(
    feature_summary$n_samples,
    feature_summary$n_raw_features,
    feature_summary$n_retained_features,
    feature_summary$n_removed_features
  )
)

knitr::kable(summary_table, align = c("l", "r"))
```

```{r metadata-summary}
meta$sample_type <- dplyr::case_when(
  meta$Class == "C" ~ "Cow",
  meta$Class == "S" ~ "Sheep",
  TRUE ~ "QC"
)

class_batch_summary <- meta |>
  dplyr::count(Batch, sample_type, name = "n_samples") |>
  dplyr::arrange(Batch, sample_type)

knitr::kable(class_batch_summary, align = c("r", "l", "r"))
```

```{r class-batch-plot}
class_batch_plot <- ggplot2::ggplot(
  class_batch_summary,
  ggplot2::aes(x = factor(Batch), y = n_samples, fill = sample_type)
) +
  ggplot2::geom_col(position = "stack", color = "#1f1f1f", linewidth = 0.25) +
  ggplot2::scale_fill_manual(values = c(Cow = "#4e79a7", Sheep = "#59a14f", QC = "#b07aa1")) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(
    title = "Public MTBLS79 sample composition by analytical batch",
    x = "Batch",
    y = "Samples",
    fill = "Class"
  ) +
  ggplot2::theme_minimal(base_size = 12)

class_batch_plot
```

## 4. Missing-value handling overview

```{r missingness-summary}
missingness_summary <- tibble::tibble(
  Stage = c("Before 80% filtering", "After 80% filtering", "After missForest imputation"),
  Missing_entries = c(
    sum(is.na(raw_data)),
    sum(is.na(public_data$filtered_data)),
    sum(is.na(imputed_data))
  )
)

knitr::kable(missingness_summary, align = c("l", "r"))
```

## 5. Saved processed files

```{r saved-files}
saved_files <- tibble::tibble(
  File = basename(public_data$output_paths),
  Path = unname(public_data$output_paths)
)

knitr::kable(saved_files, align = c("l", "l"))
```

The outputs are the manuscript-aligned processed feature matrix and metadata table used for the public MTBLS79 benchmark. The raw spreadsheet is public, but it is not bundled in this repository so that users retrieve it directly from MetaboLights.
