---
title: "BatchCorrMetabolomics Set 1: correction benchmark"
author: "Reproducible public-data benchmark"
date: "2026-07-22"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    number_sections: true
params:
  rerun_benchmark: false
---

```{r setup, include=FALSE}
find_repo_root <- function(start = getwd()) {
  here <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(here, "scripts", "run_batchcorr_set1_benchmark.R"))) return(here)
    parent <- dirname(here)
    if (identical(parent, here)) stop("Could not locate the repository root.")
    here <- parent
  }
}

repo_root <- find_repo_root()
knitr::opts_knit$set(root.dir = repo_root)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", dpi = 120)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(jsonlite)
  library(knitr)
  library(tidyr)
})
source(file.path(repo_root, "scripts", "plot_theme.R"))

result_dir <- file.path(repo_root, "results", "batchcorr_set1")
processed_dir <- file.path(repo_root, "data", "public", "batchcorr_set1", "processed")
figure_data_dir <- file.path(result_dir, "figures", "source_data")

if (isTRUE(params$rerun_benchmark)) {
  r_exec <- file.path(R.home(), "bin", "exec", "R")
  env <- c(
    paste0("R_HOME=", R.home()),
    paste0("R_SHARE_DIR=", R.home("share")),
    paste0("R_INCLUDE_DIR=", R.home("include")),
    paste0("R_DOC_DIR=", R.home("doc"))
  )
  status <- system2(
    r_exec,
    args = c("--vanilla", "--no-echo", "--file=scripts/run_batchcorr_set1_benchmark.R", "--args", "--force"),
    stdout = "",
    stderr = "",
    env = env
  )
  if (!identical(status, 0L)) stop("Benchmark rerun failed with status ", status, ". See results/batchcorr_set1/analysis_log.txt.")
}

required_results <- c(
  "analysis_complete.json", "method_attempts.csv", "method_metrics.csv",
  "method_parameters.csv", "hidden_qc_method_protocol.csv"
)
missing_results <- required_results[!file.exists(file.path(result_dir, required_results))]
if (length(missing_results)) stop("Missing benchmark output(s): ", paste(missing_results, collapse = ", "))

theme_set(theme_minimal(base_size = 11))
method_order <- winn_method_order()
method_palette <- winn_method_palette()

metadata <- read.csv(file.path(processed_dir, "BatchCorr_set1_metadata.csv"), stringsAsFactors = FALSE)
attempts <- read.csv(file.path(result_dir, "method_attempts.csv"), check.names = FALSE, stringsAsFactors = FALSE)
metrics <- read.csv(file.path(result_dir, "method_metrics.csv"), check.names = FALSE, stringsAsFactors = FALSE)
parameters <- read.csv(file.path(result_dir, "method_parameters.csv"), check.names = FALSE, stringsAsFactors = FALSE)
protocol <- read.csv(file.path(result_dir, "hidden_qc_method_protocol.csv"), check.names = FALSE, stringsAsFactors = FALSE)
completion <- read_json(file.path(result_dir, "analysis_complete.json"), simplifyVector = TRUE)
preprocessing <- read_json(file.path(processed_dir, "BatchCorr_set1_preprocessing_summary.json"), simplifyVector = TRUE)
```

# Suitability decision

**Verdict: suitable with limitations.** BatchCorrMetabolomics Set 1 supplies a minimally processed but explicitly uncorrected negative-ion LC–MS matrix, exact acquisition sequence (`SeqNr`), ten batch labels, pooled references (`SCode == "ref"`), and accession identities with genuine biological replicates. The formal criterion-level decision is in `reports/batchcorr_set1_suitability.md` and `.csv`.

The earlier oral-cavity candidate was audited first and stopped as unsuitable: its small public files do not include an injection-level feature-intensity matrix. Its raw archives were inventoried but not downloaded. No correction benchmark was run for that dataset.

# Provenance and preprocessing

The source is `set.1` plus `set.1.Y` from `rwehrens/BatchCorrMetabolomics`, pinned to commit `e0c7668140e206dcdae9afa602dd2e1b337ac4f6`. The 761 × 567 deposited matrix has 206,376 missing values (47.83%) and no zeros. Features observed in at least 80% of all injections were retained, leaving 199 features with 3.11% missingness; the remaining values were imputed by missForest with seed 42 and 100 trees.

The deposited values are documented as already log-scaled, but the log base is not documented. Imputation was therefore performed on that scale. A reversible `expm1` representation supplies nonnegative multiplicative inputs to the correction wrappers, and every evaluation uses `log1p`, returning exactly to the deposited/imputed scale. Feature IDs preserve original column positions because the package object has no m/z, retention time, or feature names.

```{r preprocessing-summary}
batch_summary <- metadata |>
  mutate(type = ifelse(is_qc, "QC", "Study")) |>
  count(batch_order_index, batch, type, name = "n") |>
  pivot_wider(names_from = type, values_from = n, values_fill = 0) |>
  mutate(Total = QC + Study) |>
  arrange(batch_order_index) |>
  select(Batch = batch, Total, Study, QC)
kable(batch_summary, caption = "Analytical batches in true acquisition order.")
```

There are 710 study injections, 51 pooled references, and 357 accession IDs. Of these, 351 accessions have two injections, one has three, and five are singletons; 346 accessions span more than one batch. Replicate analyses use only genuine repeated accession measurements.

```{r dataset-design, fig.width=11, fig.height=5}
design <- read.csv(file.path(figure_data_dir, "dataset_design.csv"), stringsAsFactors = FALSE)
ggplot(design, aes(run_order, batch_order_index, color = sample_type, shape = sample_type)) +
  geom_point(alpha = 0.75, size = 1.4) +
  scale_y_continuous(breaks = sort(unique(design$batch_order_index)),
                     labels = unique(design$batch[order(design$batch_order_index)])) +
  scale_color_manual(values = c(Study = "#9CA3AF", QC = "#D55E00")) +
  labs(title = "BatchCorr Set 1 acquisition design", subtitle = "True SeqNr order; pooled references in orange",
       x = "Global acquisition sequence number", y = "Analytical batch", color = NULL, shape = NULL)
```

# Hidden-QC protocol

Ten interior references—one per batch, selected with fresh seed 20260809—were fixed as the common holdout. This set has no overlap with the superseded QC-RLSC audit holdout. The first and last reference in every batch remain training controls, and each batch retains four or five training references. Before every method call, held-out references are relabelled as ordinary study/test samples. Their identities and values are used only after correction for evaluation, never for fitting, parameter selection, or WiNN auto-selection.

```{r holdout-tables}
holdout <- read.csv(file.path(result_dir, "qc_holdout_ids.csv"), stringsAsFactors = FALSE)
kable(holdout, caption = "Common held-out reference injections.")
kable(protocol, caption = "Method-facing information firewall.")
```

# Method execution and selected parameters

All nine requested variants were attempted. Tunable competitor candidates were ranked using accession association, accession weighted-PC variance, genuine replicate agreement, feature repeatability, and batch weighted-PC variance. Held-out QC metrics were excluded from tuning. Complete grids and rankings are under `results/batchcorr_set1/tuning_candidates/`.

QC-RLSC is fixed rather than tuned because only four or five training QCs remain per batch. A Set 1-specific wrapper performs subtractive, degree-1, full-span LOESS separately within each supplied batch on the exact `log1p` analysis scale, preserves returned sample IDs explicitly, and performs batch shifting on that log scale. This replaces a withdrawn implementation that allowed `qcrlscR::qc.rlsc.wrap()` to reorder batches and silently ignore the nominal span grid. The complete audit is in `reports/batchcorr_set1_qc_rlsc_correction.md`.

```{r methods}
kable(attempts, digits = 3, caption = "Execution status and runtime, including tuning.")
kable(parameters, caption = "Selected parameters and control sets.")
```

```{r method-coverage-note, results='asis'}
rfsc_features <- metrics$retained_features[metrics$method == "QC-RFSC"]
rfsc_removed <- 199L - rfsc_features
cat(sprintf("QC-RFSC completed but its selected package-level CV rule removed %d features, so it returns %d rather than 199 features. No other method-specific sample or feature reduction occurred; this loss is explicit in all coverage tables.", rfsc_removed, rfsc_features))
```

# Technical correction and biological preservation

Technical metrics use the ten hidden references or study-only within-batch profiles as appropriate. Biological metrics use the 710 non-QC injections. Accession is the biological identity endpoint; no pseudo-replicates were constructed.

```{r headline-metrics}
headline <- metrics |>
  transmute(
    Method = method,
    Features = retained_features,
    Samples = output_samples,
    `QC CV mean (%)` = heldout_qc_cv_mean,
    `GAM deviance` = run_order_gam_mean,
    `Ljung-Box significant (%)` = 100 * ljung_significant_proportion,
    `Batch weighted-PC R2` = batch_pc_r2,
    `Accession weighted-PC R2` = accession_pc_r2,
    `Associated features` = accession_associated_features,
    `Replicate Pearson` = replicate_pearson_median,
    `Replicate ICC(A,1)` = replicate_icc_median,
    `Runtime (s)` = runtime_sec
  )
kable(headline, digits = 3, caption = "Primary technical and biological metrics.")
```

Associated-feature counts use a feature-wise one-way accession F-test on the deposited/imputed log analysis scale, followed by Benjamini–Hochberg correction at adjusted p < 0.05. Eta-squared is saved for every feature and method. A larger associated-feature count alone is not interpreted as proof of better correction; weighted-PC identity variance and genuine replicate agreement provide complementary preservation checks.

```{r method-comparison, fig.width=12, fig.height=7}
comparison <- read.csv(file.path(figure_data_dir, "method_comparison.csv"), stringsAsFactors = FALSE)
comparison$method <- factor(comparison$method, levels = method_order)
ggplot(comparison, aes(method, value, fill = method)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = method_palette, guide = "none") +
  labs(title = "Technical correction and biological preservation", x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 32, hjust = 1))
```

```{r pca, fig.width=12, fig.height=5.5}
pca <- read.csv(file.path(figure_data_dir, "raw_vs_winn_pca.csv"), stringsAsFactors = FALSE)
ggplot(pca, aes(PC1, PC2, color = batch, group = interaction(method, accession_id))) +
  geom_line(color = "#D1D5DB", alpha = 0.25, linewidth = 0.25) +
  geom_point(alpha = 0.68, size = 1.1) +
  facet_wrap(~method, scales = "free") +
  labs(title = "Raw versus WiNN auto (QC)", subtitle = "Lines connect genuine accession replicates; color denotes batch",
       color = "Batch")
```

```{r association-change, fig.width=11, fig.height=5}
assoc_change <- read.csv(file.path(figure_data_dir, "accession_association_change.csv"), stringsAsFactors = FALSE)
assoc_change$method <- factor(assoc_change$method, levels = method_order)
ggplot(assoc_change, aes(method, change_from_raw, fill = method)) +
  geom_col() + geom_hline(yintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = method_palette, guide = "none") +
  labs(title = "Change in accession-associated features relative to Raw",
       subtitle = "BH-adjusted p < 0.05; count alone is not a correction-quality score",
       x = NULL, y = "Associated-feature count change") +
  theme(axis.text.x = element_text(angle = 32, hjust = 1))
```

```{r qc-trends, fig.width=12, fig.height=10}
qc_trends <- read.csv(file.path(figure_data_dir, "representative_qc_trends.csv"), stringsAsFactors = FALSE)
ggplot(qc_trends, aes(run_order, log_intensity, color = qc_role, shape = qc_role)) +
  geom_line(aes(group = interaction(method, feature_id, batch)), color = "#D1D5DB", linewidth = 0.25) +
  geom_point(size = 1.2) +
  facet_grid(feature_id ~ method, scales = "free_y") +
  scale_color_manual(values = c(Training = "#0072B2", `Held out` = "#D55E00")) +
  labs(title = "Representative pooled-reference trajectories",
       subtitle = "Features selected reproducibly at fixed Raw QC-CV quantiles",
       x = "Global acquisition sequence number", y = "Deposited/imputed log analysis scale",
       color = NULL, shape = NULL)
```

# Reproducibility and validation

```{r validation}
matrix_files <- list.files(file.path(result_dir, "method_outputs"), pattern = "\\.rds$", full.names = TRUE)
matrix_validation <- bind_rows(lapply(matrix_files, function(path) {
  mat <- readRDS(path)
  data.frame(
    file = basename(path), features = nrow(mat), samples = ncol(mat),
    metadata_aligned = identical(colnames(mat), metadata$sample_id[metadata$sample_id %in% colnames(mat)]),
    finite = all(is.finite(mat)), stringsAsFactors = FALSE
  )
}))
checks <- data.frame(
  check = c(
    "All nine method labels attempted exactly once",
    "No held-out QC IDs supplied to any method",
    "No tuning score uses held-out QC values",
    "Every saved method matrix aligns to metadata",
    "Every saved method matrix is finite",
    "Genuine accession IDs used for replicate metrics"
  ),
  passed = c(
    identical(attempts$method, method_order) && nrow(attempts) == 9,
    all(!protocol$supplied_holdout_qc_ids),
    all(!protocol$tuning_uses_holdout_values),
    all(matrix_validation$metadata_aligned),
    all(matrix_validation$finite),
    all(c("accession_id", "biological_replicate") %in% names(metadata))
  )
)
kable(checks, caption = "Notebook validation checks.")
kable(matrix_validation, caption = "Saved matrix dimensions and alignment.")
```

All numerical source data for the figures are saved under `results/batchcorr_set1/figures/source_data/`. Complete feature-level QC CV, GAM, Ljung–Box, accession association, and replicate tables are saved alongside the cached method matrices, selected parameters, warnings/errors, analysis log, and `sessionInfo.txt`.

```{r session-info}
sessionInfo()
```
