---
title: "Public Dataset Benchmark (MTBLS79): Technical Correction Comparison"
author: "Tanmay Tanna <ttanna@ethz.ch>"
output:
  rmarkdown::html_notebook:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 180
)

options(
  width = 130,
  dplyr.summarise.inform = FALSE,
  pdf.useDingbats = FALSE
)

resolve_repo_root <- function() {
  candidates <- c(".", "..")
  for (cand in candidates) {
    if (
      file.exists(file.path(cand, "LICENSE")) &&
      dir.exists(file.path(cand, "data")) &&
      dir.exists(file.path(cand, "notebooks"))
    ) {
      return(normalizePath(cand, mustWork = TRUE))
    }
  }
  stop("Could not locate repository root.")
}

repo_root <- resolve_repo_root()
```

This notebook benchmarks technical correction methods on the public MTBLS79 dataset. It tunes each tunable method on a feature subset, runs full-data correction, evaluates held-out QC-CV, residual run-order GAM deviance, per-feature ICC(A,1), per-sample Pearson correlation, and reports batch variance, sample-type variance, and the number of sample-type-associated metabolites.

## 1. Dependencies

```{r dependencies}
ensure_cran_package <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to load required CRAN package: ", pkg)
  }
}

ensure_bioc_package <- function(pkg) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to load required Bioconductor package: ", pkg)
  }
}

ensure_github_package <- function(pkg, repo) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    ensure_cran_package("remotes")
    remotes::install_github(repo, upgrade = "never")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to install required GitHub package: ", pkg, " from ", repo)
  }
}

cran_pkgs <- c(
  "ggplot2", "dplyr", "tidyr", "purrr", "scales", "glue", "tibble",
  "RColorBrewer", "qcrlscR", "pmartR", "rlang", "mgcv"
)
invisible(lapply(cran_pkgs, ensure_cran_package))

ensure_bioc_package("sva")
ensure_bioc_package("statTarget")
ensure_bioc_package("limma")
ensure_github_package("malbacR", "pmartR/malbacR")
ensure_github_package("TIGERr", "HAN-Siyu/TIGER")
ensure_github_package("winn", "ratschlab/winn")

library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(scales)
library(glue)
library(tibble)
library(RColorBrewer)
library(sva)
library(statTarget)
library(qcrlscR)
library(pmartR)
library(malbacR)
library(TIGERr)
library(limma)
library(winn)

if (requireNamespace("BiocParallel", quietly = TRUE)) {
  BiocParallel::register(BiocParallel::SerialParam(progressbar = FALSE), default = TRUE)
}

options(mc.cores = 1)

set.seed(20260306)
```

## 2. Styling And Global Objects

```{r style}
plot_dir <- file.path(repo_root, "notebooks", ".cache", "public_data_comparison")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

method_order_all <- c(
  "Raw",
  "ComBat",
  "QC-RLSC",
  "QC-RFSC",
  "TIGER",
  "SERRF",
  "WINN auto (QC)",
  "WINN auto-batch (QC)",
  "WINN default (no QC)"
)

external_methods <- c(
  "Raw",
  "ComBat",
  "QC-RLSC",
  "QC-RFSC",
  "TIGER",
  "SERRF",
  "WINN auto (QC)"
)

winn_methods <- c(
  "WINN auto (QC)",
  "WINN auto-batch (QC)",
  "WINN default (no QC)"
)

method_palette <- c(
  "Raw" = "#4D4D4D",
  "ComBat" = "#0072B2",
  "QC-RLSC" = "#009E73",
  "QC-RFSC" = "#E69F00",
  "TIGER" = "#D55E00",
  "SERRF" = "#CC79A7",
  "WINN auto (QC)" = "#56B4E9",
  "WINN auto-batch (QC)" = "#F0E442",
  "WINN default (no QC)" = "#000000"
)

theme_publication <- function() {
  theme_minimal(base_size = 12, base_family = "Helvetica") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "#E5E7EB", linewidth = 0.4, linetype = "dotted"),
      axis.title = element_text(color = "#111827", face = "bold", size = 12),
      axis.text = element_text(color = "#374151", size = 10),
      axis.ticks = element_line(color = "#9CA3AF"),
      strip.text = element_text(face = "bold", color = "#111827", size = 11),
      strip.background = element_rect(fill = "#F3F4F6", color = NA),
      plot.title = element_text(face = "bold", size = 13, color = "#111827", margin = margin(b = 5), lineheight = 1.05),
      plot.subtitle = element_text(size = 10.2, color = "#4B5563", margin = margin(b = 10), lineheight = 1.05),
      legend.title = element_text(face = "bold", color = "#111827"),
      legend.text = element_text(color = "#374151"),
      legend.position = "top",
      text = element_text(color = "#111827")
    )
}

format_value_label <- function(x, digits = 2, integer_threshold = 1e-10) {
  ifelse(abs(x - round(x)) < integer_threshold, sprintf("%.0f", round(x)), sprintf(paste0("%.", digits, "f"), x))
}

build_contiguous_run_segments <- function(meta_df, sample_id_col = "sample_id", order_col = "run_order", batch_col = "batch") {
  md <- meta_df[order(meta_df[[order_col]], na.last = TRUE), , drop = FALSE]
  if (nrow(md) == 0L) {
    md$segment_index <- integer(0)
    md$segment_id <- character(0)
    return(md)
  }

  batch_vals <- as.character(md[[batch_col]])
  segment_index <- cumsum(c(TRUE, batch_vals[-1] != batch_vals[-length(batch_vals)]))

  md$segment_index <- segment_index
  md$segment_id <- paste0("batch_", batch_vals, "_segment_", sprintf("%03d", segment_index))
  md
}

fit_run_order_gam_explained <- function(series, run_order, min_obs = 6L, k_max = 6L) {
  ok <- is.finite(series) & is.finite(run_order)
  n_obs <- sum(ok)

  if (n_obs < min_obs) {
    return(c(explained = NA_real_, n_obs = n_obs, k = NA_real_))
  }

  y <- as.numeric(series[ok])
  x <- as.numeric(run_order[ok])
  n_unique_x <- length(unique(x))

  if (n_unique_x < 4L) {
    return(c(explained = NA_real_, n_obs = n_obs, k = NA_real_))
  }

  if (stats::sd(y) <= .Machine$double.eps) {
    return(c(explained = 0, n_obs = n_obs, k = NA_real_))
  }

  k_use <- min(as.integer(k_max), n_unique_x - 1L, max(3L, floor(n_obs / 2L)))
  if (!is.finite(k_use) || k_use < 3L) {
    return(c(explained = NA_real_, n_obs = n_obs, k = NA_real_))
  }

  fit <- tryCatch(
    mgcv::gam(y ~ s(x, bs = "cr", k = k_use), method = "REML"),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(c(explained = NA_real_, n_obs = n_obs, k = as.numeric(k_use)))
  }

  fit_summary <- tryCatch(summary(fit), error = function(e) NULL)
  if (is.null(fit_summary)) {
    return(c(explained = NA_real_, n_obs = n_obs, k = as.numeric(k_use)))
  }

  explained <- if (is.finite(fit_summary$dev.expl)) fit_summary$dev.expl else fit_summary$r.sq
  explained <- if (is.finite(explained)) min(max(as.numeric(explained), 0), 1) else NA_real_

  c(explained = explained, n_obs = n_obs, k = as.numeric(k_use))
}

compute_metabolite_segment_gam <- function(
  mat,
  meta_df,
  method_label,
  sample_id_col = "sample_id",
  order_col = "run_order",
  batch_col = "batch",
  transform_fun = log1p,
  min_obs = 6L,
  k_max = 6L
) {
  md <- meta_df[meta_df[[sample_id_col]] %in% colnames(mat), , drop = FALSE]
  md <- build_contiguous_run_segments(md, sample_id_col = sample_id_col, order_col = order_col, batch_col = batch_col)

  if (nrow(md) == 0L || nrow(mat) == 0L) {
    return(tibble::tibble(
      method = character(),
      metabolite = character(),
      batch = character(),
      segment_id = character(),
      n_obs = integer(),
      k = numeric(),
      explained = numeric()
    ))
  }

  seg_ids <- unique(md$segment_id)
  transformed <- transform_fun(mat[, md[[sample_id_col]], drop = FALSE])

  dplyr::bind_rows(lapply(seg_ids, function(seg_id) {
    seg_md <- md[md$segment_id == seg_id, , drop = FALSE]
    seg_sample_ids <- seg_md[[sample_id_col]]
    seg_mat <- transformed[, seg_sample_ids, drop = FALSE]

    gam_stats <- t(vapply(
      seq_len(nrow(seg_mat)),
      function(i) fit_run_order_gam_explained(seg_mat[i, ], seg_md[[order_col]], min_obs = min_obs, k_max = k_max),
      numeric(3)
    ))

    tibble::tibble(
      method = method_label,
      metabolite = rownames(seg_mat),
      batch = as.character(seg_md[[batch_col]][1]),
      segment_id = seg_id,
      n_obs = as.integer(gam_stats[, "n_obs"]),
      k = as.numeric(gam_stats[, "k"]),
      explained = as.numeric(gam_stats[, "explained"])
    )
  }))
}

summarise_metabolite_segment_gam <- function(df, digits = 4) {
  df |>
    dplyr::group_by(method) |>
    dplyr::summarise(
      n_pairs = sum(is.finite(explained)),
      mean_explained = mean(explained, na.rm = TRUE),
      median_explained = stats::median(explained, na.rm = TRUE),
      sd_explained = stats::sd(explained, na.rm = TRUE),
      mean_sd = ifelse(
        is.finite(mean_explained),
        sprintf(paste0("%.", digits, "f \u00b1 %.", digits, "f"), mean_explained, ifelse(is.finite(sd_explained), sd_explained, NA_real_)),
        NA_character_
      ),
      .groups = "drop"
    ) |>
    arrange(method)
}

ljung_box_series_metrics <- function(series, max_lag = 10L, min_obs = 8L) {
  x <- as.numeric(series)
  x <- x[is.finite(x)]
  n_obs <- length(x)
  lag_use <- min(as.integer(max_lag), max(1L, floor(n_obs / 3L)), n_obs - 1L)

  if (n_obs < min_obs || lag_use < 1L || isTRUE(stats::sd(x, na.rm = TRUE) < .Machine$double.eps)) {
    return(c(lb_stat = NA_real_, p_value = NA_real_, lag = as.numeric(lag_use), n_obs = as.numeric(n_obs)))
  }

  lb <- tryCatch(
    stats::Box.test(x, lag = lag_use, type = "Ljung-Box"),
    error = function(e) NULL
  )

  if (is.null(lb)) {
    return(c(lb_stat = NA_real_, p_value = NA_real_, lag = as.numeric(lag_use), n_obs = as.numeric(n_obs)))
  }

  c(
    lb_stat = as.numeric(lb$statistic),
    p_value = as.numeric(lb$p.value),
    lag = as.numeric(lag_use),
    n_obs = as.numeric(n_obs)
  )
}

compute_metabolite_segment_ljung_box <- function(
  mat,
  meta_df,
  method_label,
  sample_id_col = "sample_id",
  order_col = "run_order",
  batch_col = "batch",
  transform_fun = log1p,
  min_obs = 8L,
  max_lag = 10L
) {
  md <- meta_df[meta_df[[sample_id_col]] %in% colnames(mat), , drop = FALSE]
  md <- build_contiguous_run_segments(md, sample_id_col = sample_id_col, order_col = order_col, batch_col = batch_col)

  if (nrow(md) == 0L || nrow(mat) == 0L) {
    return(tibble::tibble(
      method = character(),
      metabolite = character(),
      batch = character(),
      segment_id = character(),
      n_obs = integer(),
      lag = numeric(),
      lb_stat = numeric(),
      p_value = numeric()
    ))
  }

  seg_ids <- unique(md$segment_id)
  transformed <- transform_fun(mat[, md[[sample_id_col]], drop = FALSE])

  dplyr::bind_rows(lapply(seg_ids, function(seg_id) {
    seg_md <- md[md$segment_id == seg_id, , drop = FALSE]
    seg_sample_ids <- seg_md[[sample_id_col]]
    seg_mat <- transformed[, seg_sample_ids, drop = FALSE]

    lb_stats <- t(vapply(
      seq_len(nrow(seg_mat)),
      function(i) ljung_box_series_metrics(seg_mat[i, ], max_lag = max_lag, min_obs = min_obs),
      numeric(4)
    ))

    tibble::tibble(
      method = method_label,
      metabolite = rownames(seg_mat),
      batch = as.character(seg_md[[batch_col]][1]),
      segment_id = seg_id,
      n_obs = as.integer(lb_stats[, "n_obs"]),
      lag = as.numeric(lb_stats[, "lag"]),
      lb_stat = as.numeric(lb_stats[, "lb_stat"]),
      p_value = as.numeric(lb_stats[, "p_value"])
    )
  }))
}

annotate_metabolite_segment_ljung_box <- function(df, alpha = 0.05, adjust_method = "BH") {
  df |>
    dplyr::group_by(method) |>
    dplyr::mutate(
      p_adj = stats::p.adjust(p_value, method = adjust_method),
      is_autocorrelated = is.finite(p_adj) & (p_adj < alpha)
    ) |>
    dplyr::ungroup()
}

summarise_metabolite_segment_ljung_box <- function(df) {
  df |>
    dplyr::group_by(method) |>
    dplyr::summarise(
      n_tested = sum(is.finite(p_value)),
      n_autocorrelated = sum(is_autocorrelated, na.rm = TRUE),
      prop_autocorrelated = ifelse(n_tested > 0, n_autocorrelated / n_tested, NA_real_),
      median_lb_stat = stats::median(lb_stat, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(method)
}
```

## 3. Data Load And Validation

```{r data}
processed_dir <- file.path(repo_root, "data", "public", "processed")
meta_path <- file.path(processed_dir, "MTBLS79_metadata.csv")
data_path <- file.path(processed_dir, "MTBLS79_imputed_data.csv")

if (!file.exists(meta_path) || !file.exists(data_path)) {
  stop(
    "Processed public files were not found in ", processed_dir,
    ". Generate them with scripts/preprocess_mtbls79_public_data.R before running this notebook."
  )
}

meta <- read.csv(meta_path, row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
data_df <- read.csv(data_path, row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)

meta$sample_id <- rownames(meta)
meta$batch <- as.integer(meta$Batch)
meta$run_order <- seq_len(nrow(meta))
meta$class <- meta$Class
meta$sample_rep <- as.integer(meta$Sample_Rep)
meta$sample_type <- case_when(
  meta$class == "C" ~ "Cow",
  meta$class == "S" ~ "Sheep",
  TRUE ~ "QC"
)

if (!identical(rownames(data_df), meta$sample_id)) {
  common <- intersect(rownames(data_df), meta$sample_id)
  if (length(common) < 3L) stop("Metadata and data have insufficient overlap on sample IDs.")
  meta <- meta[match(common, meta$sample_id), , drop = FALSE]
  data_df <- data_df[common, , drop = FALSE]
}

x <- t(as.matrix(data_df))
storage.mode(x) <- "double"
colnames(x) <- meta$sample_id

qc_idx <- which(meta$class == "QC")
non_qc_idx <- which(meta$class != "QC")

non_qc_meta <- meta[non_qc_idx, , drop = FALSE] |>
  arrange(sample_id)

cohort_summary <- tibble(
  Metric = c(
    "Total samples",
    "QC samples",
    "Non-QC samples",
    "Features",
    "Batches",
    "Cow samples (C)",
    "Sheep samples (S)",
    "Unique non-QC Sample_Rep IDs"
  ),
  Value = c(
    ncol(x),
    length(qc_idx),
    length(non_qc_idx),
    nrow(x),
    dplyr::n_distinct(meta$batch),
    sum(meta$class == "C"),
    sum(meta$class == "S"),
    dplyr::n_distinct(non_qc_meta$sample_rep)
  )
)

knitr::kable(cohort_summary, align = c("l", "r"))
```

## 4. Method Helpers

```{r method-helpers}
clip_nonnegative <- function(m) {
  m <- as.matrix(m)
  storage.mode(m) <- "double"
  m[is.infinite(m)] <- NA_real_
  pmax(m, 0)
}

subset_to_features <- function(mat, feature_ids) {
  feature_ids <- intersect(feature_ids, rownames(mat))
  mat[feature_ids, , drop = FALSE]
}

safe_cor <- function(a, b, method = "pearson") {
  suppressWarnings(cor(a, b, use = "pairwise.complete.obs", method = method))
}

calc_icc_a1_matrix <- function(ratings) {
  ratings <- as.matrix(ratings)
  ratings <- ratings[complete.cases(ratings), , drop = FALSE]

  n <- nrow(ratings)
  k <- ncol(ratings)
  if (n < 2L || k < 2L) {
    return(NA_real_)
  }

  grand_mean <- mean(ratings)
  row_means <- rowMeans(ratings)
  col_means <- colMeans(ratings)

  ss_rows <- k * sum((row_means - grand_mean)^2)
  ss_cols <- n * sum((col_means - grand_mean)^2)
  ss_total <- sum((ratings - grand_mean)^2)
  ss_error <- ss_total - ss_rows - ss_cols

  ms_rows <- ss_rows / (n - 1L)
  ms_cols <- ss_cols / (k - 1L)
  ms_error <- ss_error / ((n - 1L) * (k - 1L))

  denom <- ms_rows + (k - 1L) * ms_error + (k * (ms_cols - ms_error) / n)
  if (!is.finite(denom) || denom <= 0) {
    return(NA_real_)
  }

  (ms_rows - ms_error) / denom
}

align_meta_to_mat <- function(x_mat, meta_df = meta) {
  sample_ids <- colnames(x_mat)
  md <- meta_df[match(sample_ids, meta_df$sample_id), , drop = FALSE]

  if (anyNA(md$sample_id)) {
    stop("Sample IDs in matrix do not match metadata sample IDs.")
  }

  md
}

attach_qc_passthrough <- function(corrected_mat, raw_mat, meta_df) {
  qc_samples <- meta_df$sample_id[meta_df$class == "QC" & meta_df$sample_id %in% colnames(raw_mat)]
  if (length(qc_samples) == 0L) {
    return(corrected_mat)
  }

  feature_ids <- intersect(rownames(corrected_mat), rownames(raw_mat))
  corrected_mat <- corrected_mat[feature_ids, , drop = FALSE]
  raw_aligned <- raw_mat[feature_ids, , drop = FALSE]

  missing_qc <- setdiff(qc_samples, colnames(corrected_mat))
  if (length(missing_qc) > 0L) {
    corrected_mat <- cbind(corrected_mat, raw_aligned[, missing_qc, drop = FALSE])
  }

  output_ids <- meta_df$sample_id[meta_df$sample_id %in% colnames(corrected_mat)]
  corrected_mat[, output_ids, drop = FALSE]
}

run_method <- function(label, fn, tuning_sec = 0) {
  process_sec <- system.time({
    out <- fn()
  })[["elapsed"]]

  tuning_sec <- as.numeric(tuning_sec)[1]
  if (!is.finite(tuning_sec)) {
    tuning_sec <- 0
  }

  list(
    label = label,
    data = clip_nonnegative(out),
    tuning_sec = tuning_sec,
    process_sec = process_sec,
    elapsed_sec = tuning_sec + process_sec
  )
}

run_method_cached <- function(label, fn, tuning_sec = 0, force = FALSE, cache_version = "") {
  cache_dir <- file.path(plot_dir, "method_cache")
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

  safe_label <- gsub("[^A-Za-z0-9]+", "_", tolower(label))
  cache_suffix <- ifelse(nzchar(cache_version), paste0("_", cache_version), "")
  cache_file <- file.path(cache_dir, paste0(safe_label, cache_suffix, ".rds"))

  if (!isTRUE(force) && file.exists(cache_file)) {
    out <- readRDS(cache_file)
    out$tuning_sec <- as.numeric(tuning_sec)[1]
    if (!is.finite(out$tuning_sec)) out$tuning_sec <- 0
    if (is.null(out$process_sec) || !is.finite(out$process_sec)) {
      out$process_sec <- out$elapsed_sec
    }
    out$elapsed_sec <- out$process_sec + out$tuning_sec
    return(out)
  }

  out <- run_method(label = label, fn = fn, tuning_sec = tuning_sec)
  saveRDS(out, cache_file)
  out
}

run_silently <- function(code) {
  suppressWarnings(suppressMessages(force(code)))
}

with_no_cluster <- function(code, patch_do_parallel = FALSE) {
  parallel_ns <- asNamespace("parallel")
  original_parallel <- list(
    makeCluster = get("makeCluster", envir = parallel_ns),
    clusterExport = get("clusterExport", envir = parallel_ns),
    stopCluster = get("stopCluster", envir = parallel_ns)
  )

  if (bindingIsLocked("makeCluster", parallel_ns)) unlockBinding("makeCluster", parallel_ns)
  if (bindingIsLocked("clusterExport", parallel_ns)) unlockBinding("clusterExport", parallel_ns)
  if (bindingIsLocked("stopCluster", parallel_ns)) unlockBinding("stopCluster", parallel_ns)

  assign("makeCluster", function(spec, ...) NULL, envir = parallel_ns)
  assign("clusterExport", function(cl, ...) invisible(NULL), envir = parallel_ns)
  assign("stopCluster", function(cl, ...) invisible(NULL), envir = parallel_ns)

  lockBinding("makeCluster", parallel_ns)
  lockBinding("clusterExport", parallel_ns)
  lockBinding("stopCluster", parallel_ns)

  do_parallel_ns <- NULL
  original_register <- NULL

  if (isTRUE(patch_do_parallel) && requireNamespace("doParallel", quietly = TRUE)) {
    do_parallel_ns <- asNamespace("doParallel")
    original_register <- get("registerDoParallel", envir = do_parallel_ns)
    if (bindingIsLocked("registerDoParallel", do_parallel_ns)) unlockBinding("registerDoParallel", do_parallel_ns)
    assign("registerDoParallel", function(cl, ...) foreach::registerDoSEQ(), envir = do_parallel_ns)
    lockBinding("registerDoParallel", do_parallel_ns)
  }

  on.exit({
    if (!is.null(do_parallel_ns) && !is.null(original_register)) {
      if (bindingIsLocked("registerDoParallel", do_parallel_ns)) unlockBinding("registerDoParallel", do_parallel_ns)
      assign("registerDoParallel", original_register, envir = do_parallel_ns)
      lockBinding("registerDoParallel", do_parallel_ns)
    }

    if (bindingIsLocked("makeCluster", parallel_ns)) unlockBinding("makeCluster", parallel_ns)
    if (bindingIsLocked("clusterExport", parallel_ns)) unlockBinding("clusterExport", parallel_ns)
    if (bindingIsLocked("stopCluster", parallel_ns)) unlockBinding("stopCluster", parallel_ns)

    assign("makeCluster", original_parallel$makeCluster, envir = parallel_ns)
    assign("clusterExport", original_parallel$clusterExport, envir = parallel_ns)
    assign("stopCluster", original_parallel$stopCluster, envir = parallel_ns)

    lockBinding("makeCluster", parallel_ns)
    lockBinding("clusterExport", parallel_ns)
    lockBinding("stopCluster", parallel_ns)
  }, add = TRUE)

  force(code)
}

make_malbac_object <- function(x_mat, meta_df, jitter_eps = 0) {
  x_mat <- as.matrix(x_mat)
  if (jitter_eps > 0) {
    x_mat <- x_mat + matrix(runif(length(x_mat), min = 0, max = jitter_eps), nrow = nrow(x_mat))
  }

  safe_sample <- paste0("Sample_", seq_len(ncol(x_mat)))
  sample_map <- setNames(meta_df$sample_id, safe_sample)

  x_safe <- x_mat
  colnames(x_safe) <- safe_sample

  f_data <- data.frame(
    sampleID = safe_sample,
    sample_type = meta_df$class,
    batch = meta_df$batch,
    run_order = meta_df$run_order,
    stringsAsFactors = FALSE
  )
  rownames(f_data) <- f_data$sampleID

  e_data <- data.frame(
    feature = rownames(x_safe),
    x_safe,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

  obj <- pmartR::as.metabData(
    e_data = e_data,
    f_data = f_data,
    edata_cname = "feature",
    fdata_cname = "sampleID"
  )

  obj <- pmartR::group_designation(
    omicsData = obj,
    main_effects = "sample_type",
    batch_id = "batch"
  )

  list(obj = obj, sample_map = sample_map)
}

extract_malbac_matrix <- function(malbac_obj, sample_map) {
  e_data <- malbac_obj$e_data
  m <- as.matrix(e_data[, -1, drop = FALSE])
  rownames(m) <- e_data[, 1]

  mapped <- sample_map[colnames(m)]
  colnames(m) <- unname(mapped)
  m
}

run_combat <- function(x_mat, par_prior = TRUE) {
  x_log <- log1p(x_mat)
  y_log <- sva::ComBat(
    dat = x_log,
    batch = meta$batch,
    mod = NULL,
    par.prior = par_prior,
    prior.plots = FALSE
  )
  expm1(y_log)
}

run_qc_rlsc <- function(x_mat, span = 0.75) {
  md <- meta |>
    arrange(run_order)
  x_ord <- x_mat[, md$sample_id, drop = FALSE]
  dat <- as.data.frame(t(x_ord), check.names = FALSE)
  y <- ifelse(md$class == "QC", "QC", "Sample")
  corrected <- qcrlscR::qc.rlsc.wrap(
    dat = dat,
    cls.qc = y,
    cls.bl = factor(md$batch),
    intra = TRUE,
    opti = FALSE,
    log10 = FALSE,
    outl = FALSE,
    shift = TRUE,
    span = span
  )
  corrected <- t(as.matrix(corrected))
  rownames(corrected) <- rownames(x_ord)
  colnames(corrected) <- md$sample_id
  corrected[, meta$sample_id, drop = FALSE]
}

run_qc_rfsc <- function(x_mat, ntree = 500, coCV = 30, Frule = 0.8) {
  stat_names <- ifelse(meta$class == "QC", paste0(meta$sample_id, "_QC"), meta$sample_id)

  samPeno <- data.frame(
    sample = stat_names,
    batch = meta$batch,
    class = ifelse(meta$class == "QC", NA_integer_, 1L),
    order = meta$run_order,
    stringsAsFactors = FALSE
  )

  if (!is.na(samPeno$class[1]) || !is.na(samPeno$class[nrow(samPeno)])) {
    stop("statTarget::shiftCor requires QC samples at first and last run-order positions.")
  }

  samFile <- data.frame(name = rownames(x_mat), x_mat, check.names = FALSE)
  colnames(samFile)[-1] <- stat_names

  tmp <- tempfile("mtbls79_shiftcor_")
  dir.create(tmp)
  old <- getwd()
  setwd(tmp)
  on.exit(setwd(old), add = TRUE)

  write.csv(samPeno, "samPeno.csv", row.names = FALSE, quote = FALSE)
  write.csv(samFile, "samFile.csv", row.names = FALSE, quote = FALSE)

  run_silently(
    statTarget::shiftCor(
      samPeno = "samPeno.csv",
      samFile = "samFile.csv",
      Frule = Frule,
      MLmethod = "QCRFSC",
      ntree = ntree,
      coCV = coCV,
      plot = FALSE
    )
  )

  out_file <- file.path("statTarget", "shiftCor", "After_shiftCor", "shift_all_cor.csv")
  if (!file.exists(out_file)) stop("statTarget did not produce shift_all_cor.csv")

  corrected_df <- read.csv(out_file, check.names = FALSE, row.names = 1, stringsAsFactors = FALSE)
  if ("class" %in% rownames(corrected_df)) {
    corrected_df <- corrected_df[rownames(corrected_df) != "class", , drop = FALSE]
  }

  corrected <- apply(corrected_df, 2, as.numeric)
  rownames(corrected) <- rownames(corrected_df)
  colnames(corrected) <- sub("_QC$", "", colnames(corrected_df))
  corrected <- corrected / 1000

  corrected[, meta$sample_id, drop = FALSE]
}

run_tiger <- function(x_mat, use_injection = TRUE, include_qc_passthrough = TRUE) {
  md <- align_meta_to_mat(x_mat, meta)

  sample_df <- tibble(
    sampleID = md$sample_id,
    sampleType = md$class,
    batchID = md$batch,
    injectionOrder = md$run_order
  ) |>
    bind_cols(as.data.frame(t(x_mat), check.names = FALSE))

  train_samples <- sample_df |> filter(sampleType == "QC")
  if (nrow(train_samples) < 2L) {
    stop("TIGER requires at least two QC samples in training data.")
  }

  test_samples <- sample_df |> filter(sampleType != "QC")

  if (!isTRUE(use_injection)) {
    train_samples <- train_samples |> select(-injectionOrder)
    test_samples <- test_samples |> select(-injectionOrder)
  }

  corrected <- run_silently(
    TIGERr::run_TIGER(
      test_samples = test_samples,
      train_samples = train_samples,
      col_sampleID = "sampleID",
      col_sampleType = "sampleType",
      col_batchID = "batchID",
      col_order = if (isTRUE(use_injection)) "injectionOrder" else NULL
    )
  )

  meta_cols <- c("sampleID", "sampleType", "batchID")
  if (isTRUE(use_injection)) {
    meta_cols <- c(meta_cols, "injectionOrder")
  }

  feature_cols <- setdiff(colnames(corrected), meta_cols)
  m <- t(as.matrix(corrected[, feature_cols, drop = FALSE]))
  rownames(m) <- feature_cols
  colnames(m) <- corrected$sampleID

  if (isTRUE(include_qc_passthrough)) {
    attach_qc_passthrough(m, x_mat, md)
  } else {
    m
  }
}

run_serrf <- function(x_mat, jitter_eps = 0, include_qc_passthrough = TRUE) {
  md <- align_meta_to_mat(x_mat, meta)
  mal <- make_malbac_object(x_mat, md, jitter_eps = jitter_eps)

  out <- run_silently(
    with_no_cluster(
      malbacR::bc_serrf(
        omicsData = mal$obj,
        sampletype_cname = "sample_type",
        test_val = "QC",
        group_cname = "sample_type"
      ),
      patch_do_parallel = TRUE
    )
  )

  corrected <- extract_malbac_matrix(out, mal$sample_map)
  corrected <- corrected[, md$sample_id[md$sample_id %in% colnames(corrected)], drop = FALSE]

  if (!isTRUE(include_qc_passthrough)) {
    return(corrected)
  }

  attach_qc_passthrough(corrected, x_mat, md)
}

run_winn_core <- function(x_mat, batch_vec, control_ids = NULL, parameters_mode = "auto") {
  md <- align_meta_to_mat(x_mat, meta) |>
    arrange(run_order)
  x_ord <- x_mat[, md$sample_id, drop = FALSE]
  batch_arg <- if (is.null(batch_vec)) NULL else md$batch

  control_idx <- NULL
  if (!is.null(control_ids)) {
    control_idx <- match(control_ids, md$sample_id, nomatch = 0L)
    control_idx <- control_idx[control_idx > 0L]
  }

  corrected <- winn::winn(
    data = x_ord,
    batch = batch_arg,
    run_order = md$run_order,
    control_samples = control_idx,
    parameters = parameters_mode
  )

  corrected[, colnames(x_mat), drop = FALSE]
}

run_winn_mode1_auto_qc <- function(x_mat) {
  run_winn_core(
    x_mat = x_mat,
    batch_vec = meta$batch,
    control_ids = meta$sample_id[qc_idx],
    parameters_mode = "auto"
  )
}

run_winn_mode2_auto_batch_qc <- function(x_mat) {
  run_winn_core(
    x_mat = x_mat,
    batch_vec = NULL,
    control_ids = meta$sample_id[qc_idx],
    parameters_mode = "auto"
  )
}

run_winn_mode3_default_no_qc <- function(x_mat) {
  run_winn_core(
    x_mat = x_mat,
    batch_vec = meta$batch,
    control_ids = NULL,
    parameters_mode = "fixed"
  )
}
```

## 5. Assessment Helpers

```{r assessment-helpers}
weighted_pc_r2 <- function(mat, target, n_pcs = 5L) {
  target <- as.factor(target)
  x_log <- log1p(mat)
  x_t <- t(x_log)

  keep <- apply(x_t, 2, function(v) {
    vv <- v[is.finite(v)]
    if (length(vv) < 2L) return(FALSE)
    stats::sd(vv) > 0
  })
  x_t <- x_t[, keep, drop = FALSE]

  if (nrow(x_t) < 3L || ncol(x_t) < 2L) {
    return(NA_real_)
  }

  pca <- tryCatch(
    prcomp(x_t, center = TRUE, scale. = TRUE),
    error = function(e) NULL
  )
  if (is.null(pca)) {
    return(NA_real_)
  }

  pcs <- min(as.integer(n_pcs), ncol(pca$x))
  if (pcs < 1L) {
    return(NA_real_)
  }

  scores <- pca$x[, seq_len(pcs), drop = FALSE]
  var_exp <- (pca$sdev^2) / sum(pca$sdev^2)
  var_exp <- var_exp[seq_len(pcs)]

  r2 <- vapply(seq_len(pcs), function(i) {
    summary(lm(scores[, i] ~ target))$r.squared
  }, numeric(1))

  sum(r2 * var_exp, na.rm = TRUE) / sum(var_exp, na.rm = TRUE)
}

collapse_by_sample_rep <- function(mat, md) {
  split_idx <- split(seq_len(ncol(mat)), md$sample_rep)
  out <- vapply(split_idx, function(idx) rowMeans(log1p(mat[, idx, drop = FALSE]), na.rm = TRUE), numeric(nrow(mat)))
  if (is.null(dim(out))) {
    out <- matrix(out, nrow = nrow(mat), ncol = 1L)
  }
  rownames(out) <- rownames(mat)

  rep_type <- vapply(split_idx, function(idx) {
    types <- md$class[idx]
    names(which.max(table(types)))
  }, character(1))

  list(mat = out, rep_type = rep_type)
}

count_bio_metabolites <- function(mat, md, fdr_cutoff = 0.05, effect_cutoff = 0.2) {
  collapsed <- collapse_by_sample_rep(mat, md)
  cm <- collapsed$mat
  rep_type <- collapsed$rep_type

  c_idx <- which(rep_type == "C")
  s_idx <- which(rep_type == "S")

  pvals <- apply(cm, 1, function(v) {
    if (length(c_idx) < 2 || length(s_idx) < 2) return(NA_real_)
    tryCatch(t.test(v[c_idx], v[s_idx])$p.value, error = function(e) NA_real_)
  })

  effect <- rowMeans(cm[, c_idx, drop = FALSE], na.rm = TRUE) -
    rowMeans(cm[, s_idx, drop = FALSE], na.rm = TRUE)

  padj <- p.adjust(pvals, method = "BH")
  sig <- is.finite(padj) & padj < fdr_cutoff & is.finite(effect) & abs(effect) >= effect_cutoff

  list(
    n_sig = sum(sig, na.rm = TRUE),
    p_adj = padj,
    effect = effect,
    sig = sig
  )
}

sample_repeat_metrics <- function(mat, md) {
  split_idx <- split(seq_len(ncol(mat)), md$sample_rep)
  x_log <- log1p(mat)

  out <- lapply(names(split_idx), function(rep_id) {
    idx <- split_idx[[rep_id]]
    if (length(idx) < 2L) {
      return(NULL)
    }

    corr <- suppressWarnings(cor(x_log[, idx, drop = FALSE], use = "pairwise.complete.obs"))
    vals <- corr[lower.tri(corr)]
    icc_val <- calc_icc_a1_matrix(x_log[, idx, drop = FALSE])

    tibble(
      sample_rep = as.integer(rep_id),
      pearson_median = median(vals, na.rm = TRUE),
      icc_a1 = icc_val
    )
  })

  bind_rows(out)
}

make_batch_pair_index <- function(md) {
  batches <- sort(unique(md$batch))
  combn(batches, 2, simplify = FALSE, FUN = function(bpair) {
    b1 <- bpair[1]
    b2 <- bpair[2]

    idx1 <- which(md$batch == b1)
    idx2 <- which(md$batch == b2)

    rep1 <- md$sample_rep[idx1]
    rep2 <- md$sample_rep[idx2]

    common_rep <- intersect(rep1, rep2)
    if (length(common_rep) < 3L) {
      return(NULL)
    }

    aligned1 <- idx1[match(common_rep, rep1)]
    aligned2 <- idx2[match(common_rep, rep2)]

    list(batch1 = b1, batch2 = b2, idx1 = aligned1, idx2 = aligned2)
  }) |> purrr::compact()
}

feature_batch_consistency <- function(mat, pair_index, cor_method = "pearson") {
  x_log <- log1p(mat)

  apply(x_log, 1, function(v) {
    vals <- vapply(pair_index, function(p) {
      safe_cor(v[p$idx1], v[p$idx2], method = cor_method)
    }, numeric(1))

    median(vals, na.rm = TRUE)
  })
}

feature_batch_icc <- function(mat, pair_index) {
  x_log <- log1p(mat)

  apply(x_log, 1, function(v) {
    vals <- vapply(pair_index, function(p) {
      calc_icc_a1_matrix(cbind(v[p$idx1], v[p$idx2]))
    }, numeric(1))

    median(vals, na.rm = TRUE)
  })
}

coverage_penalized_feature_corr <- function(method_corr, raw_corr, full_feature_ids) {
  full_feature_ids <- intersect(full_feature_ids, names(raw_corr))
  penalized <- raw_corr[full_feature_ids]
  names(penalized) <- full_feature_ids

  common <- intersect(names(method_corr), full_feature_ids)
  penalized[common] <- method_corr[common]
  penalized
}

evaluate_method <- function(mat, md, pair_index) {
  bio <- count_bio_metabolites(mat, md)
  rep_sim <- sample_repeat_metrics(mat, md)
  feat_cons_pearson <- feature_batch_consistency(mat, pair_index, cor_method = "pearson")
  feat_cons_icc <- feature_batch_icc(mat, pair_index)

  list(
    metabolites_retained = nrow(mat),
    bio_n = bio$n_sig,
    type_pc_r2 = weighted_pc_r2(mat, md$class, n_pcs = 5L),
    batch_pc_r2 = weighted_pc_r2(mat, md$batch, n_pcs = 5L),
    repeat_df = rep_sim,
    feature_consistency_pearson = feat_cons_pearson,
    feature_consistency_icc = feat_cons_icc,
    repeat_median = median(rep_sim$pearson_median, na.rm = TRUE),
    repeat_mean = mean(rep_sim$pearson_median, na.rm = TRUE),
    repeat_sd = sd(rep_sim$pearson_median, na.rm = TRUE),
    repeat_icc_median = median(rep_sim$icc_a1, na.rm = TRUE),
    repeat_icc_mean = mean(rep_sim$icc_a1, na.rm = TRUE),
    repeat_icc_sd = sd(rep_sim$icc_a1, na.rm = TRUE),
    feature_consistency_median_pearson = median(feat_cons_pearson, na.rm = TRUE),
    feature_consistency_mean_pearson = mean(feat_cons_pearson, na.rm = TRUE),
    feature_consistency_sd_pearson = sd(feat_cons_pearson, na.rm = TRUE),
    feature_consistency_median_icc = median(feat_cons_icc, na.rm = TRUE),
    feature_consistency_mean_icc = mean(feat_cons_icc, na.rm = TRUE),
    feature_consistency_sd_icc = sd(feat_cons_icc, na.rm = TRUE)
  )
}
```

## 6. Parameter Tuning On A Small Feature Subset

```{r tuning, cache=TRUE}
set.seed(20260306)
tune_n_features <- min(220L, nrow(x))
tune_features <- sample(rownames(x), size = tune_n_features, replace = FALSE)
x_tune <- subset_to_features(x, tune_features)
tuning_log_file <- file.path(plot_dir, "tuning_progress.log")
cat("", file = tuning_log_file)
tuning_timeout_sec <- 600

log_tuning <- function(msg) {
  cat(sprintf("[%s] %s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), msg), file = tuning_log_file, append = TRUE)
}

run_with_timeout <- function(code, timeout_sec = 600) {
  setTimeLimit(elapsed = timeout_sec, transient = TRUE)
  on.exit(setTimeLimit(cpu = Inf, elapsed = Inf, transient = FALSE), add = TRUE)
  force(code)
}

tune_rank <- function(x, higher_better = TRUE) {
  x <- as.numeric(x)
  if (higher_better) {
    x[!is.finite(x)] <- -Inf
    rank(x, ties.method = "average")
  } else {
    x[!is.finite(x)] <- Inf
    rank(-x, ties.method = "average")
  }
}

score_tuning_table <- function(df) {
  if (nrow(df) <= 1L) {
    df$score <- 1
    return(df)
  }

  df$score <-
    tune_rank(df$bio_n, higher_better = TRUE) +
    tune_rank(df$type_pc_r2, higher_better = TRUE) +
    tune_rank(df$repeat_median, higher_better = TRUE) +
    tune_rank(df$feature_consistency_median_pearson_penalized, higher_better = TRUE) +
    tune_rank(df$batch_pc_r2, higher_better = FALSE)

  df
}

full_features_tune <- rownames(x_tune)

tune_generic <- function(grid_df, run_fn, method_label = "method", timeout_sec = 600) {
  rows <- vector("list", nrow(grid_df))

  for (i in seq_len(nrow(grid_df))) {
    p <- as.list(grid_df[i, , drop = FALSE])
    log_tuning(sprintf("%s candidate %d/%d started", method_label, i, nrow(grid_df)))

    m <- tryCatch(
      run_with_timeout(run_fn(p), timeout_sec = timeout_sec),
      error = function(e) {
        log_tuning(sprintf("%s candidate %d failed: %s", method_label, i, conditionMessage(e)))
        NULL
      }
    )

    if (is.null(m) || !is.matrix(m)) {
      log_tuning(sprintf("%s candidate %d skipped (no valid matrix)", method_label, i))
      next
    }

    common_samples <- non_qc_meta$sample_id[non_qc_meta$sample_id %in% colnames(m)]
    if (length(common_samples) < 20L) {
      log_tuning(sprintf("%s candidate %d skipped (insufficient common samples)", method_label, i))
      next
    }

    md <- non_qc_meta[match(common_samples, non_qc_meta$sample_id), , drop = FALSE]
    m_eval <- m[, common_samples, drop = FALSE]
    pair_index_md <- make_batch_pair_index(md)

    ev <- evaluate_method(m_eval, md, pair_index_md)
    raw_corr_md <- feature_batch_consistency(
      x_tune[, common_samples, drop = FALSE],
      pair_index_md,
      cor_method = "pearson"
    )
    feat_corr_penalized <- coverage_penalized_feature_corr(
      method_corr = ev$feature_consistency_pearson,
      raw_corr = raw_corr_md,
      full_feature_ids = full_features_tune
    )

    rows[[i]] <- cbind(
      grid_df[i, , drop = FALSE],
      bio_n = ev$bio_n,
      type_pc_r2 = ev$type_pc_r2,
      batch_pc_r2 = ev$batch_pc_r2,
      repeat_median = ev$repeat_median,
      feature_consistency_median_pearson = ev$feature_consistency_median_pearson,
      feature_consistency_median_pearson_penalized = median(feat_corr_penalized, na.rm = TRUE)
    )

    log_tuning(sprintf("%s candidate %d completed", method_label, i))
  }

  out <- bind_rows(rows)
  if (nrow(out) == 0L) {
    fallback <- grid_df[1, , drop = FALSE]
    fallback$bio_n <- NA_real_
    fallback$type_pc_r2 <- NA_real_
    fallback$batch_pc_r2 <- NA_real_
    fallback$repeat_median <- NA_real_
    fallback$feature_consistency_median_pearson <- NA_real_
    fallback$feature_consistency_median_pearson_penalized <- NA_real_
    fallback$score <- NA_real_
    return(fallback)
  }

  score_tuning_table(out) |>
    arrange(desc(score))
}

run_tuning_timed <- function(grid_df, run_fn, method_label) {
  elapsed <- system.time({
    tbl <- tune_generic(grid_df, run_fn, method_label = method_label, timeout_sec = tuning_timeout_sec)
  })[["elapsed"]]
  list(table = tbl, elapsed_sec = elapsed)
}

combat_grid <- tibble(par_prior = c(TRUE, FALSE))
qcrlsc_grid <- tibble(span = c(0.30, 0.50, 0.70, 0.90))
qcrfsc_grid <- tidyr::expand_grid(
  ntree = c(200L, 500L),
  coCV = c(20, 30),
  Frule = c(0.8)
)
tiger_grid <- tibble(use_injection = c(TRUE, FALSE))
serrf_grid <- tibble(jitter_eps = c(0, 1e-6, 1e-5))

combat_tuned <- run_tuning_timed(combat_grid, function(p) {
  run_combat(x_tune, par_prior = as.logical(p$par_prior))
}, method_label = "ComBat")
combat_tuning <- combat_tuned$table

qcrlsc_tuned <- run_tuning_timed(qcrlsc_grid, function(p) {
  run_qc_rlsc(x_tune, span = as.numeric(p$span))
}, method_label = "QC-RLSC")
qcrlsc_tuning <- qcrlsc_tuned$table

qcrfsc_tuned <- run_tuning_timed(qcrfsc_grid, function(p) {
  run_qc_rfsc(
    x_tune,
    ntree = as.integer(p$ntree),
    coCV = as.numeric(p$coCV),
    Frule = as.numeric(p$Frule)
  )
}, method_label = "QC-RFSC")
qcrfsc_tuning <- qcrfsc_tuned$table

tiger_tuned <- run_tuning_timed(tiger_grid, function(p) {
  run_tiger(x_tune, use_injection = as.logical(p$use_injection))
}, method_label = "TIGER")
tiger_tuning <- tiger_tuned$table

serrf_tuned <- run_tuning_timed(serrf_grid, function(p) {
  run_serrf(x_tune, jitter_eps = as.numeric(p$jitter_eps))
}, method_label = "SERRF")
serrf_tuning <- serrf_tuned$table

tuning_runtime_sec <- c(
  "ComBat" = combat_tuned$elapsed_sec,
  "QC-RLSC" = qcrlsc_tuned$elapsed_sec,
  "QC-RFSC" = qcrfsc_tuned$elapsed_sec,
  "TIGER" = tiger_tuned$elapsed_sec,
  "SERRF" = serrf_tuned$elapsed_sec
)

best_params <- list(
  combat = combat_tuning[1, , drop = FALSE],
  qcrlsc = qcrlsc_tuning[1, , drop = FALSE],
  qcrfsc = qcrfsc_tuning[1, , drop = FALSE],
  tiger = tiger_tuning[1, , drop = FALSE],
  serrf = serrf_tuning[1, , drop = FALSE]
)

selected_params <- bind_rows(
  best_params$combat |> mutate(method = "ComBat"),
  best_params$qcrlsc |> mutate(method = "QC-RLSC"),
  best_params$qcrfsc |> mutate(method = "QC-RFSC"),
  best_params$tiger |> mutate(method = "TIGER"),
  best_params$serrf |> mutate(method = "SERRF")
) |>
  relocate(method)
```

## 7. Run Full Methods With Tuned Parameters

```{r run-methods, cache=TRUE}
tuning_time_for <- function(label) {
  val <- tuning_runtime_sec[[label]]
  if (is.null(val) || !is.finite(val)) {
    return(0)
  }
  as.numeric(val)
}

method_results <- list(
  run_method_cached("Raw", function() x, tuning_sec = 0),
  run_method_cached("ComBat", function() {
    run_combat(x, par_prior = as.logical(best_params$combat$par_prior[1]))
  }, tuning_sec = tuning_time_for("ComBat")),
  run_method_cached("QC-RLSC", function() {
    run_qc_rlsc(x, span = as.numeric(best_params$qcrlsc$span[1]))
  }, tuning_sec = tuning_time_for("QC-RLSC")),
  run_method_cached("QC-RFSC", function() {
    run_qc_rfsc(
      x,
      ntree = as.integer(best_params$qcrfsc$ntree[1]),
      coCV = as.numeric(best_params$qcrfsc$coCV[1]),
      Frule = as.numeric(best_params$qcrfsc$Frule[1])
    )
  }, tuning_sec = tuning_time_for("QC-RFSC")),
  run_method_cached("TIGER", function() {
    run_tiger(x, use_injection = as.logical(best_params$tiger$use_injection[1]))
  }, tuning_sec = tuning_time_for("TIGER"), cache_version = "tigerr_default_v1"),
  run_method_cached("SERRF", function() {
    run_serrf(x, jitter_eps = as.numeric(best_params$serrf$jitter_eps[1]))
  }, tuning_sec = tuning_time_for("SERRF")),
  run_method_cached("WINN auto (QC)", function() {
    run_winn_mode1_auto_qc(x)
  }, tuning_sec = 0),
  run_method_cached("WINN auto-batch (QC)", function() {
    run_winn_mode2_auto_batch_qc(x)
  }, tuning_sec = 0),
  run_method_cached("WINN default (no QC)", function() {
    run_winn_mode3_default_no_qc(x)
  }, tuning_sec = 0)
)

method_results <- lapply(method_results, function(res) {
  if (res$label %in% c("TIGER", "SERRF")) {
    res$data <- attach_qc_passthrough(res$data, x, meta)
  }
  res
})

method_mats <- setNames(lapply(method_results, `[[`, "data"), vapply(method_results, `[[`, character(1), "label"))
```

## 8. Evaluate Methods

```{r evaluate, cache=TRUE}
align_method_set <- function(method_list, required_samples) {
  common_samples <- Reduce(intersect, lapply(method_list, colnames))
  common_samples <- required_samples[required_samples %in% common_samples]

  aligned <- lapply(method_list, function(m) {
    m[, common_samples, drop = FALSE]
  })

  list(methods = aligned, samples = common_samples)
}

external_set <- align_method_set(
  method_list = method_mats[external_methods],
  required_samples = non_qc_meta$sample_id
)

winn_set <- align_method_set(
  method_list = method_mats[winn_methods],
  required_samples = non_qc_meta$sample_id
)

external_meta <- non_qc_meta[match(external_set$samples, non_qc_meta$sample_id), , drop = FALSE]
winn_meta <- non_qc_meta[match(winn_set$samples, non_qc_meta$sample_id), , drop = FALSE]

external_pair_index <- make_batch_pair_index(external_meta)
winn_pair_index <- make_batch_pair_index(winn_meta)

external_eval <- lapply(external_set$methods, evaluate_method, md = external_meta, pair_index = external_pair_index)
winn_eval <- lapply(winn_set$methods, evaluate_method, md = winn_meta, pair_index = winn_pair_index)

common_feature_panel_all_methods <- Reduce(intersect, lapply(method_mats, rownames))
full_feature_panel <- rownames(method_mats[["Raw"]])

add_penalized_pearson <- function(eval_list, raw_corr, full_feature_ids) {
  lapply(eval_list, function(ev) {
    penalized <- coverage_penalized_feature_corr(
      method_corr = ev$feature_consistency_pearson,
      raw_corr = raw_corr,
      full_feature_ids = full_feature_ids
    )
    ev$feature_consistency_pearson_penalized <- penalized
    ev$feature_consistency_median_pearson_penalized <- median(penalized, na.rm = TRUE)
    ev$feature_consistency_mean_pearson_penalized <- mean(penalized, na.rm = TRUE)
    ev$feature_consistency_sd_pearson_penalized <- sd(penalized, na.rm = TRUE)
    ev
  })
}

add_penalized_icc <- function(eval_list, raw_icc, full_feature_ids) {
  lapply(eval_list, function(ev) {
    penalized <- coverage_penalized_feature_corr(
      method_corr = ev$feature_consistency_icc,
      raw_corr = raw_icc,
      full_feature_ids = full_feature_ids
    )
    ev$feature_consistency_icc_penalized <- penalized
    ev$feature_consistency_median_icc_penalized <- median(penalized, na.rm = TRUE)
    ev$feature_consistency_mean_icc_penalized <- mean(penalized, na.rm = TRUE)
    ev$feature_consistency_sd_icc_penalized <- sd(penalized, na.rm = TRUE)
    ev
  })
}

add_common_retained_pearson <- function(eval_list, method_list, sample_ids, pair_index, common_feature_ids) {
  lapply(names(eval_list), function(label) {
    ev <- eval_list[[label]]
    mat <- method_list[[label]]
    common_features <- intersect(common_feature_ids, rownames(mat))

    if (length(common_features) < 2L) {
      ev$feature_consistency_common_pearson <- setNames(numeric(0), character(0))
      ev$feature_consistency_median_common_pearson <- NA_real_
      ev$feature_consistency_mean_common_pearson <- NA_real_
      ev$feature_consistency_sd_common_pearson <- NA_real_
      return(ev)
    }

    mat_common <- mat[common_features, sample_ids, drop = FALSE]
    corr_common <- feature_batch_consistency(mat_common, pair_index, cor_method = "pearson")
    ev$feature_consistency_common_pearson <- corr_common
    ev$feature_consistency_median_common_pearson <- median(corr_common, na.rm = TRUE)
    ev$feature_consistency_mean_common_pearson <- mean(corr_common, na.rm = TRUE)
    ev$feature_consistency_sd_common_pearson <- sd(corr_common, na.rm = TRUE)
    ev
  }) |> setNames(names(eval_list))
}

add_common_retained_icc <- function(eval_list, method_list, pair_index, common_feature_ids) {
  lapply(names(eval_list), function(label) {
    ev <- eval_list[[label]]
    mat <- method_list[[label]]
    common_features <- intersect(common_feature_ids, rownames(mat))

    if (length(common_features) < 2L) {
      ev$feature_consistency_common_icc <- setNames(numeric(0), character(0))
      ev$feature_consistency_median_common_icc <- NA_real_
      ev$feature_consistency_mean_common_icc <- NA_real_
      ev$feature_consistency_sd_common_icc <- NA_real_
      return(ev)
    }

    mat_common <- mat[common_features, , drop = FALSE]
    icc_common <- feature_batch_icc(mat_common, pair_index)
    ev$feature_consistency_common_icc <- icc_common
    ev$feature_consistency_median_common_icc <- median(icc_common, na.rm = TRUE)
    ev$feature_consistency_mean_common_icc <- mean(icc_common, na.rm = TRUE)
    ev$feature_consistency_sd_common_icc <- sd(icc_common, na.rm = TRUE)
    ev
  }) |> setNames(names(eval_list))
}

raw_corr_external <- external_eval[["Raw"]]$feature_consistency_pearson
raw_corr_winn <- feature_batch_consistency(
  method_mats[["Raw"]][, winn_set$samples, drop = FALSE],
  winn_pair_index,
  cor_method = "pearson"
)
raw_icc_external <- external_eval[["Raw"]]$feature_consistency_icc
raw_icc_winn <- feature_batch_icc(
  method_mats[["Raw"]][, winn_set$samples, drop = FALSE],
  winn_pair_index
)

external_eval <- add_penalized_pearson(external_eval, raw_corr_external, full_feature_panel)
winn_eval <- add_penalized_pearson(winn_eval, raw_corr_winn, full_feature_panel)
external_eval <- add_penalized_icc(external_eval, raw_icc_external, full_feature_panel)
winn_eval <- add_penalized_icc(winn_eval, raw_icc_winn, full_feature_panel)
external_eval <- add_common_retained_pearson(
  eval_list = external_eval,
  method_list = external_set$methods,
  sample_ids = external_set$samples,
  pair_index = external_pair_index,
  common_feature_ids = common_feature_panel_all_methods
)
winn_eval <- add_common_retained_pearson(
  eval_list = winn_eval,
  method_list = winn_set$methods,
  sample_ids = winn_set$samples,
  pair_index = winn_pair_index,
  common_feature_ids = common_feature_panel_all_methods
)
external_eval <- add_common_retained_icc(
  eval_list = external_eval,
  method_list = external_set$methods,
  pair_index = external_pair_index,
  common_feature_ids = common_feature_panel_all_methods
)
winn_eval <- add_common_retained_icc(
  eval_list = winn_eval,
  method_list = winn_set$methods,
  pair_index = winn_pair_index,
  common_feature_ids = common_feature_panel_all_methods
)

run_order_segment_meta <- meta |>
  filter(class != "QC", sample_id %in% colnames(x)) |>
  arrange(run_order)

run_order_gam_full_all <- bind_rows(lapply(method_order_all, function(label) {
  compute_metabolite_segment_gam(
    method_mats[[label]],
    run_order_segment_meta,
    method_label = label,
    transform_fun = log1p,
    min_obs = 6L,
    k_max = 5L
  )
})) |>
  filter(is.finite(explained)) |>
  mutate(method = factor(method, levels = method_order_all))

run_order_gam_summary_all <- summarise_metabolite_segment_gam(run_order_gam_full_all) |>
  mutate(method = factor(method, levels = method_order_all)) |>
  arrange(method)

ljungbox_full_all <- bind_rows(lapply(method_order_all, function(label) {
  compute_metabolite_segment_ljung_box(
    method_mats[[label]],
    run_order_segment_meta,
    method_label = label,
    transform_fun = log1p,
    min_obs = 8L,
    max_lag = 10L
  )
})) |>
  annotate_metabolite_segment_ljung_box(alpha = 0.05, adjust_method = "BH") |>
  mutate(method = factor(method, levels = method_order_all))

ljungbox_summary_all <- summarise_metabolite_segment_ljung_box(ljungbox_full_all) |>
  mutate(method = factor(method, levels = method_order_all)) |>
  arrange(method)

summarize_eval <- function(eval_list) {
  tibble(
    method = names(eval_list),
    metabolites_retained = vapply(eval_list, function(e) e$metabolites_retained, numeric(1)),
    bio_metabolites = vapply(eval_list, function(e) e$bio_n, numeric(1)),
    sample_type_pc_r2 = vapply(eval_list, function(e) e$type_pc_r2, numeric(1)),
    batch_pc_r2 = vapply(eval_list, function(e) e$batch_pc_r2, numeric(1)),
    repeat_similarity_median = vapply(eval_list, function(e) e$repeat_median, numeric(1)),
    repeat_icc_median = vapply(eval_list, function(e) e$repeat_icc_median, numeric(1)),
    feature_batch_corr_median_pearson_common = vapply(eval_list, function(e) e$feature_consistency_median_common_pearson, numeric(1)),
    feature_batch_corr_median_pearson_penalized = vapply(eval_list, function(e) e$feature_consistency_median_pearson_penalized, numeric(1)),
    feature_batch_icc_median_common = vapply(eval_list, function(e) e$feature_consistency_median_common_icc, numeric(1)),
    feature_batch_icc_median_penalized = vapply(eval_list, function(e) e$feature_consistency_median_icc_penalized, numeric(1))
  ) |>
    left_join(
      run_order_gam_summary_all |>
        transmute(
          method,
          residual_run_order_gam_mean = mean_explained,
          residual_run_order_gam_median = median_explained,
          residual_run_order_gam_sd = sd_explained
        ),
      by = "method"
    ) |>
    left_join(
      ljungbox_summary_all |>
        transmute(
          method,
          ljungbox_n_autocorrelated = n_autocorrelated,
          ljungbox_prop_autocorrelated = prop_autocorrelated
        ),
      by = "method"
    )
}

external_metric_summary <- summarize_eval(external_eval) |>
  transmute(
    method = as.character(method),
    sample_type_associated_metabolites = bio_metabolites,
    sample_type_variance = sample_type_pc_r2,
    batch_variance = batch_pc_r2,
    per_sample_pearson = repeat_similarity_median,
    per_feature_icc = feature_batch_icc_median_penalized,
    residual_run_order_gam = residual_run_order_gam_mean,
    autocorrelated_metabolite_plate_count = ljungbox_n_autocorrelated
  ) |>
  mutate(method = factor(method, levels = external_methods)) |>
  arrange(method)

winn_metric_summary <- summarize_eval(winn_eval) |>
  transmute(
    method = as.character(method),
    sample_type_associated_metabolites = bio_metabolites,
    sample_type_variance = sample_type_pc_r2,
    batch_variance = batch_pc_r2,
    per_sample_pearson = repeat_similarity_median,
    per_feature_icc = feature_batch_icc_median_penalized,
    residual_run_order_gam = residual_run_order_gam_mean,
    autocorrelated_metabolite_plate_count = ljungbox_n_autocorrelated
  ) |>
  mutate(method = factor(method, levels = winn_methods)) |>
  arrange(method)

metric_summary_plot <- bind_rows(
  external_metric_summary |> mutate(method = as.character(method)),
  winn_metric_summary |>
    mutate(method = as.character(method)) |>
    filter(method %in% c("WINN auto-batch (QC)", "WINN default (no QC)"))
) |>
  mutate(method = factor(method, levels = method_order_all)) |>
  arrange(method)
```

## 9. Held-Out QC-CV

```{r qccv, cache=TRUE}
cv_percent <- function(mat, sample_ids) {
  sample_ids <- intersect(sample_ids, colnames(mat))
  if (length(sample_ids) < 3L || nrow(mat) < 2L) {
    return(setNames(rep(NA_real_, nrow(mat)), rownames(mat)))
  }

  sub <- mat[, sample_ids, drop = FALSE]
  mean_vec <- rowMeans(sub, na.rm = TRUE)
  sd_vec <- apply(sub, 1, sd, na.rm = TRUE)
  denom <- pmax(abs(mean_vec), .Machine$double.eps)
  out <- 100 * sd_vec / denom
  names(out) <- rownames(sub)
  out
}

run_hidden_qc_cached <- function(label, fn, cache_tag = "", force = FALSE) {
  cache_dir <- file.path(plot_dir, "method_cache_qc_holdout")
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

  safe_label <- gsub("[^A-Za-z0-9]+", "_", tolower(label))
  tag <- ifelse(nzchar(cache_tag), paste0("_", cache_tag), "")
  cache_file <- file.path(cache_dir, paste0(safe_label, tag, ".rds"))

  if (!isTRUE(force) && file.exists(cache_file)) {
    return(readRDS(cache_file))
  }

  out <- clip_nonnegative(fn())
  saveRDS(out, cache_file)
  out
}

run_qc_rlsc_with_controls <- function(x_mat, control_ids, span = 0.75) {
  md <- align_meta_to_mat(x_mat, meta) |>
    arrange(run_order)
  x_ord <- x_mat[, md$sample_id, drop = FALSE]
  dat <- as.data.frame(t(x_ord), check.names = FALSE)
  y <- ifelse(md$sample_id %in% control_ids, "QC", "Sample")
  corrected <- qcrlscR::qc.rlsc.wrap(
    dat = dat,
    cls.qc = y,
    cls.bl = factor(md$batch),
    intra = TRUE,
    opti = FALSE,
    log10 = FALSE,
    outl = FALSE,
    shift = TRUE,
    span = span
  )
  corrected <- t(as.matrix(corrected))
  rownames(corrected) <- rownames(x_ord)
  colnames(corrected) <- md$sample_id
  corrected[, meta$sample_id, drop = FALSE]
}

run_qc_rfsc_with_controls <- function(x_mat, control_ids, ntree = 500, coCV = 30, Frule = 0.8) {
  md <- align_meta_to_mat(x_mat, meta)
  qc_flag <- md$sample_id %in% control_ids
  stat_names <- ifelse(qc_flag, paste0(md$sample_id, "_QC"), md$sample_id)

  samPeno <- data.frame(
    sample = stat_names,
    batch = md$batch,
    class = ifelse(qc_flag, NA_integer_, 1L),
    order = md$run_order,
    stringsAsFactors = FALSE
  )

  if (!is.na(samPeno$class[1]) || !is.na(samPeno$class[nrow(samPeno)])) {
    stop("statTarget::shiftCor requires QC samples at first and last run-order positions.")
  }

  samFile <- data.frame(name = rownames(x_mat), x_mat, check.names = FALSE)
  colnames(samFile)[-1] <- stat_names

  tmp <- tempfile("mtbls79_shiftcor_holdout_")
  dir.create(tmp)
  old <- getwd()
  setwd(tmp)
  on.exit(setwd(old), add = TRUE)

  write.csv(samPeno, "samPeno.csv", row.names = FALSE, quote = FALSE)
  write.csv(samFile, "samFile.csv", row.names = FALSE, quote = FALSE)

  run_silently(
    statTarget::shiftCor(
      samPeno = "samPeno.csv",
      samFile = "samFile.csv",
      Frule = Frule,
      MLmethod = "QCRFSC",
      ntree = ntree,
      coCV = coCV,
      plot = FALSE
    )
  )

  out_file <- file.path("statTarget", "shiftCor", "After_shiftCor", "shift_all_cor.csv")
  if (!file.exists(out_file)) stop("statTarget did not produce shift_all_cor.csv")

  corrected_df <- read.csv(out_file, check.names = FALSE, row.names = 1, stringsAsFactors = FALSE)
  if ("class" %in% rownames(corrected_df)) {
    corrected_df <- corrected_df[rownames(corrected_df) != "class", , drop = FALSE]
  }

  corrected <- apply(corrected_df, 2, as.numeric)
  rownames(corrected) <- rownames(corrected_df)
  colnames(corrected) <- sub("_QC$", "", colnames(corrected_df))
  corrected <- corrected / 1000

  corrected[, md$sample_id, drop = FALSE]
}

run_tiger_with_holdout_qc <- function(
  x_mat,
  train_qc_ids,
  test_qc_ids,
  use_injection = TRUE,
  mtry_percent = 0.4,
  nodesize_percent = 0.4,
  ntree = 5,
  parallel_cores = 2
) {
  md <- align_meta_to_mat(x_mat, meta)

  train_qc_ids <- intersect(train_qc_ids, md$sample_id[md$class == "QC"])
  test_qc_ids <- intersect(test_qc_ids, md$sample_id[md$class == "QC"])
  if (length(train_qc_ids) < 2L) stop("TIGER QC-CV requires at least 2 training QCs.")
  if (length(test_qc_ids) < 1L) stop("TIGER QC-CV requires at least 1 test QC.")

  sample_df <- tibble(
    sampleID = md$sample_id,
    sampleType = ifelse(md$sample_id %in% train_qc_ids, "QC", "Sample"),
    batchID = md$batch,
    injectionOrder = md$run_order
  ) |>
    bind_cols(as.data.frame(t(x_mat), check.names = FALSE))

  train_samples <- sample_df |> filter(sampleID %in% train_qc_ids)
  test_samples <- sample_df |> filter(!(sampleID %in% train_qc_ids))

  if (!isTRUE(use_injection)) {
    train_samples <- train_samples |> select(-injectionOrder)
    test_samples <- test_samples |> select(-injectionOrder)
  }

  corrected <- run_silently(
    TIGERr::run_TIGER(
      test_samples = test_samples,
      train_samples = train_samples,
      col_sampleID = "sampleID",
      col_sampleType = "sampleType",
      col_batchID = "batchID",
      col_order = if (isTRUE(use_injection)) "injectionOrder" else NULL,
      mtry_percent = mtry_percent,
      nodesize_percent = nodesize_percent,
      ntree = ntree,
      parallel.cores = parallel_cores
    )
  )

  meta_cols <- c("sampleID", "sampleType", "batchID")
  if (isTRUE(use_injection)) meta_cols <- c(meta_cols, "injectionOrder")

  feature_cols <- setdiff(colnames(corrected), meta_cols)
  m <- t(as.matrix(corrected[, feature_cols, drop = FALSE]))
  rownames(m) <- feature_cols
  colnames(m) <- corrected$sampleID

  m[, intersect(test_qc_ids, colnames(m)), drop = FALSE]
}

run_serrf_with_holdout_qc <- function(
  x_mat,
  train_qc_ids,
  test_qc_ids,
  jitter_eps = 0,
  include_nonqc_samples = FALSE
) {
  md <- align_meta_to_mat(x_mat, meta)

  if (!isTRUE(include_nonqc_samples)) {
    keep_ids <- md$sample_id[md$class == "QC"]
    md <- md[match(keep_ids, md$sample_id), , drop = FALSE]
    x_mat <- x_mat[, keep_ids, drop = FALSE]
  }

  train_qc_ids <- intersect(train_qc_ids, md$sample_id[md$class == "QC"])
  test_qc_ids <- intersect(test_qc_ids, md$sample_id[md$class == "QC"])
  if (length(train_qc_ids) < 2L) stop("SERRF QC-CV requires at least 2 training QCs.")
  if (length(test_qc_ids) < 1L) stop("SERRF QC-CV requires at least 1 test QC.")

  md_serrf <- md
  md_serrf$class <- as.character(md_serrf$class)
  md_serrf$class[md_serrf$sample_id %in% train_qc_ids] <- "QC"
  md_serrf$class[!(md_serrf$sample_id %in% train_qc_ids)] <- "Sample"

  jitter_candidates <- unique(c(
    as.numeric(jitter_eps),
    if (isTRUE(jitter_eps <= 0)) c(1e-6, 1e-4, 1e-3, 1e-2) else c(as.numeric(jitter_eps) * 10, 1e-3, 1e-2)
  ))

  out <- NULL
  last_err <- NULL
  for (jit in jitter_candidates) {
    mal <- make_malbac_object(x_mat, md_serrf, jitter_eps = jit)
    out_try <- tryCatch(
      run_silently(
        with_no_cluster(
          malbacR::bc_serrf(
            omicsData = mal$obj,
            sampletype_cname = "sample_type",
            test_val = "QC",
            group_cname = "sample_type"
          ),
          patch_do_parallel = TRUE
        )
      ),
      error = function(e) e
    )

    if (!inherits(out_try, "error")) {
      out <- out_try
      break
    }
    last_err <- out_try
  }

  if (is.null(out)) stop(last_err)

  m <- extract_malbac_matrix(out, mal$sample_map)
  m[, intersect(test_qc_ids, colnames(m)), drop = FALSE]
}

run_winn_with_controls <- function(x_mat, control_ids, auto_batch = FALSE) {
  run_winn_core(
    x_mat = x_mat,
    batch_vec = if (isTRUE(auto_batch)) NULL else meta$batch,
    control_ids = control_ids,
    parameters_mode = "auto"
  )
}

qc_meta <- meta |>
  filter(class == "QC") |>
  arrange(run_order)

if (nrow(qc_meta) < 12L) {
  stop("At least 12 QC samples are required for QC holdout evaluation.")
}

sample_holdout_qcs <- function(qc_df, n_holdout = 10L, min_train_per_batch = 3L, seed = 20260311) {
  qc_df <- qc_df |>
    mutate(batch = as.integer(batch)) |>
    arrange(batch, run_order) |>
    group_by(batch) |>
    mutate(
      qc_rank_within_batch = row_number(),
      n_qc_batch = n(),
      eligible_holdout = qc_rank_within_batch > 1L & qc_rank_within_batch < n_qc_batch
    ) |>
    ungroup()

  batch_caps <- qc_df |>
    group_by(batch) |>
    summarise(
      n_qc_total = n(),
      n_eligible = sum(eligible_holdout),
      .groups = "drop"
    ) |>
    mutate(
      max_holdout = pmax(
        0L,
        pmin(
          n_qc_total - 2L,
          n_qc_total - as.integer(min_train_per_batch)
        )
      )
    )

  if (sum(batch_caps$max_holdout) < n_holdout) {
    stop("Cannot hold out requested number of interior QCs while preserving minimum train QCs per batch.")
  }

  eligible_batches <- batch_caps$batch[batch_caps$n_eligible > 0L & batch_caps$max_holdout > 0L]
  if (length(eligible_batches) > n_holdout) {
    stop("More batches with eligible QCs than requested hold-out size.")
  }

  selected <- character(0)
  selected_by_batch <- setNames(integer(nrow(batch_caps)), as.character(batch_caps$batch))
  set.seed(seed)

  for (batch_id in eligible_batches) {
    candidates <- qc_df$sample_id[qc_df$batch == batch_id & qc_df$eligible_holdout]
    picked_id <- sample(candidates, size = 1L)
    selected <- c(selected, picked_id)
    selected_by_batch[as.character(batch_id)] <- selected_by_batch[as.character(batch_id)] + 1L
  }

  remaining_needed <- n_holdout - length(selected)
  while (remaining_needed > 0L) {
    remaining_cap <- batch_caps$max_holdout - selected_by_batch[as.character(batch_caps$batch)]
    extra_batches <- batch_caps$batch[remaining_cap > 0L]
    extra_batches <- extra_batches[vapply(extra_batches, function(b) {
      any(qc_df$batch == b & qc_df$eligible_holdout & !(qc_df$sample_id %in% selected))
    }, logical(1))]

    if (length(extra_batches) == 0L) {
      break
    }

    weights <- remaining_cap[match(extra_batches, batch_caps$batch)]
    picked_batch <- sample(extra_batches, size = 1L, prob = weights)
    candidates <- qc_df$sample_id[qc_df$batch == picked_batch & qc_df$eligible_holdout & !(qc_df$sample_id %in% selected)]
    picked_id <- sample(candidates, size = 1L)
    selected <- c(selected, picked_id)
    selected_by_batch[as.character(picked_batch)] <- selected_by_batch[as.character(picked_batch)] + 1L
    remaining_needed <- n_holdout - length(selected)
  }

  if (length(selected) < n_holdout) {
    stop("Failed to sample requested number of holdout QCs under per-batch constraints.")
  }

  sort(selected)
}

tiger_qccv_tag <- paste0(
  "inj_", as.integer(as.logical(best_params$tiger$use_injection[1])),
  "_mtry_0.4_nodesize_0.4_ntree_5_fullrun"
)
serrf_qccv_tag <- paste0(
  "jit_", formatC(as.numeric(best_params$serrf$jitter_eps[1]), format = "f", digits = 6),
  "_fullrun"
)

holdout_seed <- 20260327L
holdout_qc_ids <- sample_holdout_qcs(
  qc_df = qc_meta,
  n_holdout = 12L,
  min_train_per_batch = 3L,
  seed = holdout_seed
)
train_qc_ids <- setdiff(qc_meta$sample_id, holdout_qc_ids)

holdout_tag_num <- abs(sum(utf8ToInt(paste(holdout_qc_ids, collapse = "|"))))
holdout_tag <- paste0("holdout12_seed", holdout_seed, "_", holdout_tag_num, "_fullrun_v5")

qccv_holdout_mats <- list(
  "Raw" = method_mats[["Raw"]],
  "ComBat" = method_mats[["ComBat"]],
  "QC-RLSC" = run_hidden_qc_cached(
    "QC-RLSC",
    function() run_qc_rlsc_with_controls(
      x,
      control_ids = train_qc_ids,
      span = as.numeric(best_params$qcrlsc$span[1])
    ),
    cache_tag = paste0(
      holdout_tag, "_qcrlscsubtract_v1_",
      sprintf("span_%.3f", as.numeric(best_params$qcrlsc$span[1]))
    )
  ),
  "QC-RFSC" = run_hidden_qc_cached(
    "QC-RFSC",
    function() run_qc_rfsc_with_controls(
      x,
      control_ids = train_qc_ids,
      ntree = as.integer(best_params$qcrfsc$ntree[1]),
      coCV = as.numeric(best_params$qcrfsc$coCV[1]),
      Frule = as.numeric(best_params$qcrfsc$Frule[1])
    ),
    cache_tag = paste0(
      holdout_tag, "_",
      "nt", as.integer(best_params$qcrfsc$ntree[1]),
      "_cv", formatC(as.numeric(best_params$qcrfsc$coCV[1]), format = "f", digits = 2),
      "_fr", formatC(as.numeric(best_params$qcrfsc$Frule[1]), format = "f", digits = 2)
    )
  ),
  "TIGER" = run_hidden_qc_cached(
    "TIGER_QCCV_HOLDOUT",
    function() run_tiger_with_holdout_qc(
      x,
      train_qc_ids = train_qc_ids,
      test_qc_ids = holdout_qc_ids,
      use_injection = as.logical(best_params$tiger$use_injection[1]),
      mtry_percent = 0.4,
      nodesize_percent = 0.4,
      ntree = 5,
      parallel_cores = 2
    ),
    cache_tag = paste0(holdout_tag, "_", tiger_qccv_tag)
  ),
  "SERRF" = run_hidden_qc_cached(
    "SERRF_QCCV_HOLDOUT",
    function() run_serrf_with_holdout_qc(
      x,
      train_qc_ids = train_qc_ids,
      test_qc_ids = holdout_qc_ids,
      jitter_eps = as.numeric(best_params$serrf$jitter_eps[1]),
      include_nonqc_samples = TRUE
    ),
    cache_tag = paste0(holdout_tag, "_", serrf_qccv_tag)
  ),
  "WINN auto (QC)" = run_hidden_qc_cached(
    "WINN auto (QC)",
    function() run_winn_with_controls(x, control_ids = train_qc_ids, auto_batch = FALSE),
    cache_tag = paste0(holdout_tag, "_winnauto_v3")
  ),
  "WINN auto-batch (QC)" = run_hidden_qc_cached(
    "WINN auto-batch (QC)",
    function() run_winn_with_controls(x, control_ids = train_qc_ids, auto_batch = TRUE),
    cache_tag = paste0(holdout_tag, "_winnautobatch_v3")
  ),
  "WINN default (no QC)" = method_mats[["WINN default (no QC)"]]
)

as_cv_long <- function(method_label, mat, sample_ids) {
  vals <- cv_percent(mat, sample_ids)
  tibble(
    method = method_label,
    feature = names(vals),
    value = as.numeric(vals)
  ) |>
    filter(is.finite(value))
}

qccv_holdout_diagnostics <- bind_rows(lapply(method_order_all, function(label) {
  vals <- cv_percent(qccv_holdout_mats[[label]], holdout_qc_ids)
  tibble(
    seed = holdout_seed,
    method = label,
    n_holdout_columns_present = length(intersect(holdout_qc_ids, colnames(qccv_holdout_mats[[label]]))),
    n_features_finite = sum(is.finite(vals)),
    mean_value = mean(vals, na.rm = TRUE)
  )
}))

if (any(qccv_holdout_diagnostics$n_features_finite < 1L)) {
  bad_methods <- unique(qccv_holdout_diagnostics$method[qccv_holdout_diagnostics$n_features_finite < 1L])
  stop(glue("Hold-out QC-CV produced no finite values for: {paste(bad_methods, collapse = ', ')}"))
}

holdout_training_mode <- c(
  "Raw" = "No QC labels used; CV evaluated on held-out QC IDs",
  "ComBat" = "No QC labels used; CV evaluated on held-out QC IDs",
  "QC-RLSC" = "Held-out QCs relabelled as samples during correction",
  "QC-RFSC" = "Held-out QCs relabelled as samples during correction",
  "TIGER" = "Held-out QCs moved to the TIGER test/sample set",
  "SERRF" = "Held-out QCs relabelled as samples during correction",
  "WINN auto (QC)" = "Held-out QCs excluded from the QC control set",
  "WINN auto-batch (QC)" = "Held-out QCs excluded from the QC control set",
  "WINN default (no QC)" = "No QC labels used; CV evaluated on held-out QC IDs"
)

qccv_summary <- bind_rows(lapply(method_order_all, function(label) {
  as_cv_long(
    method_label = label,
    mat = qccv_holdout_mats[[label]],
    sample_ids = holdout_qc_ids
  )
})) |>
  mutate(method = factor(method, levels = method_order_all)) |>
  group_by(method) |>
  summarise(
    n_features = sum(is.finite(value)),
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    median_value = median(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    value_label = format_value_label(mean_value, digits = 2),
    holdout_training_mode = holdout_training_mode[as.character(method)]
  ) |>
  arrange(method)

qccv_metric_summary <- qccv_summary |>
  transmute(
    method = as.character(method),
    heldout_qc_cv = mean_value,
    heldout_qc_cv_sd = sd_value
  )

external_metric_summary <- external_metric_summary |>
  mutate(method = as.character(method)) |>
  left_join(qccv_metric_summary, by = "method") |>
  mutate(method = factor(method, levels = external_methods)) |>
  arrange(method)

winn_metric_summary <- winn_metric_summary |>
  mutate(method = as.character(method)) |>
  left_join(qccv_metric_summary, by = "method") |>
  mutate(method = factor(method, levels = winn_methods)) |>
  arrange(method)

metric_summary_plot <- metric_summary_plot |>
  mutate(method = as.character(method)) |>
  left_join(qccv_metric_summary, by = "method") |>
  mutate(method = factor(method, levels = method_order_all)) |>
  arrange(method)
```

## 10. Allowed Plots

```{r plots, fig.width=14, fig.height=8}
plot_pca <- function(method_list, md, method_levels, title, ncol) {
  df <- bind_rows(lapply(names(method_list), function(label) {
    mat <- method_list[[label]]
    x_t <- t(log1p(mat))
    keep <- apply(x_t, 2, function(v) {
      vv <- v[is.finite(v)]
      if (length(vv) < 2L) return(FALSE)
      stats::sd(vv) > 0
    })
    x_t <- x_t[, keep, drop = FALSE]

    if (ncol(x_t) < 2L) {
      pc_scores <- matrix(0, nrow = nrow(x_t), ncol = 2)
      colnames(pc_scores) <- c("PC1", "PC2")
      ve <- c(0, 0)
    } else {
      pca <- tryCatch(prcomp(x_t, center = TRUE, scale. = TRUE), error = function(e) NULL)
      if (is.null(pca)) {
        pc_scores <- matrix(0, nrow = nrow(x_t), ncol = 2)
        colnames(pc_scores) <- c("PC1", "PC2")
        ve <- c(0, 0)
      } else {
        ve_all <- 100 * (pca$sdev^2) / sum(pca$sdev^2)
        ve <- c(ve_all[1], ifelse(length(ve_all) >= 2L, ve_all[2], 0))
        pc_scores <- cbind(
          PC1 = pca$x[, 1],
          PC2 = if (ncol(pca$x) >= 2L) pca$x[, 2] else rep(0, nrow(pca$x))
        )
      }
    }

    tibble(
      sample_id = md$sample_id,
      batch = factor(md$batch),
      sample_type = factor(ifelse(md$class == "C", "Cow", "Sheep"), levels = c("Cow", "Sheep")),
      PC1 = pc_scores[, "PC1"],
      PC2 = pc_scores[, "PC2"],
      method = factor(label, levels = method_levels),
      facet_label = glue("{label}\nPC1 {round(ve[1], 1)}% | PC2 {round(ve[2], 1)}%")
    )
  }))

  batch_vals <- sort(unique(md$batch))
  batch_cols <- setNames(
    RColorBrewer::brewer.pal(max(3, length(batch_vals)), "Dark2")[seq_along(batch_vals)],
    batch_vals
  )

  ggplot(df, aes(PC1, PC2, color = batch, shape = sample_type)) +
    geom_point(size = 1.9, alpha = 0.9, stroke = 0.2) +
    facet_wrap(~ facet_label, ncol = ncol) +
    scale_color_manual(values = batch_cols, name = "Batch") +
    scale_shape_manual(values = c("Cow" = 16, "Sheep" = 17), name = "Sample type") +
    labs(title = title, x = "PC1 score", y = "PC2 score") +
    coord_equal() +
    theme_publication()
}

plot_metric_bar <- function(df, value_col, y_lab, title) {
  ggplot(df, aes(method, .data[[value_col]], fill = method)) +
    geom_col(width = 0.72, color = "#1F2937", linewidth = 0.25) +
    scale_fill_manual(values = method_palette, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(title = title, x = NULL, y = y_lab) +
    theme_publication() +
    theme(axis.text.x = element_text(angle = 28, hjust = 1))
}

method_non_qc_mats <- lapply(method_mats, function(mat) {
  mat[, non_qc_meta$sample_id, drop = FALSE]
})

plot_pca(
  method_list = method_non_qc_mats,
  md = non_qc_meta,
  method_levels = method_order_all,
  title = "MTBLS79 PCA by correction method",
  ncol = 3
)

plot_metric_bar(metric_summary_plot, "batch_variance", "Weighted PC R^2", "Residual batch variance")
plot_metric_bar(metric_summary_plot, "sample_type_variance", "Weighted PC R^2", "Sample-type variance")
plot_metric_bar(metric_summary_plot, "sample_type_associated_metabolites", "Metabolites", "Sample-type-associated metabolites")
```

## 11. Summary

```{r summary}
knitr::kable(
  selected_params |>
    select(method, any_of(c("par_prior", "span", "ntree", "coCV", "Frule", "use_injection", "jitter_eps", "score"))) |>
    mutate(across(where(is.numeric), ~ round(.x, 4))),
  caption = "Selected subset-tuning parameters."
)

knitr::kable(
  external_metric_summary |>
    mutate(across(where(is.numeric), ~ round(.x, 4))),
  caption = "External comparison summary."
)

knitr::kable(
  winn_metric_summary |>
    mutate(across(where(is.numeric), ~ round(.x, 4))),
  caption = "WINN mode comparison summary."
)

holdout_summary <- tibble(
  holdout_seed = holdout_seed,
  n_holdout_qc = length(holdout_qc_ids),
  holdout_qc_ids = paste(holdout_qc_ids, collapse = ", ")
)

knitr::kable(holdout_summary, caption = "Held-out QC split used for QC-CV evaluation.")
```
