---
title: "Sacurine / MTBLS404 human public correction benchmark"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
repo_root <- normalizePath(".", mustWork = TRUE)
if (!file.exists(file.path(repo_root, "scripts", "run_human_public_benchmark.R"))) {
  repo_root <- normalizePath("..", mustWork = TRUE)
}
result_dir <- file.path(repo_root, "results", "sacurine")
report_dir <- file.path(repo_root, "reports")
read_result <- function(name) read.csv(file.path(result_dir, name), check.names = FALSE, stringsAsFactors = FALSE)
metrics <- read_result("method_metrics.csv")
metrics_common <- read_result("method_metrics_common_panel.csv")
metrics_coverage <- read_result("method_metrics_coverage_penalized.csv")
dimensions <- read_result("method_dimensions.csv")
runtime <- read_result("method_runtime.csv")
parameters <- read_result("method_parameters.csv")
holdout <- read_result("qc_holdout_ids.csv")
training <- read_result("qc_training_ids.csv")
```

# Scope and provenance

This notebook reports the executed benchmark of the pre-correction W4M00001
Sacurine negative-ion intensity table. Only the pinned, small processed tables
and their documentation were downloaded; raw spectra were neither downloaded
nor reprocessed. Technical metrics use the deposited pre-osmolality intensity
matrix. Biological models apply the same sample-wise osmolality division after
every correction method.

```{r audit, results='asis', echo=FALSE}
cat(paste(readLines(file.path(report_dir, "sacurine_suitability.md"), warn = FALSE), collapse = "\n"))
```

# Fixed hidden-QC protocol

The eight held-out pooled QCs below were selected once with seed 20260810. They
were relabelled as ordinary samples in every method-facing metadata table and
were excluded from correction training, competitor tuning, and WiNN automatic
selection. First and last QCs within each batch remained in training.

```{r qc-split, echo=FALSE}
knitr::kable(holdout, caption = "Held-out QCs")
knitr::kable(aggregate(sample_id ~ batch, training, length), caption = "Training QCs by batch")
```

# Methods attempted

All nine prespecified variants were attempted in the same order. Candidate
selection used only training-QC behavior; phenotype labels and held-out QC
values were unavailable to tuning.

```{r methods, echo=FALSE}
knitr::kable(runtime, digits = 3, caption = "Method status and runtime")
knitr::kable(parameters, caption = "Selected or fixed parameters")
knitr::kable(dimensions, caption = "Output coverage and alignment")
```

# Primary results

Technical entries in the primary table evaluate every feature each method
retained. All biological entries use the saved common feature intersection, so
age, BMI, and gender models have identical subjects and feature denominators for
all nine methods. Any method-specific loss is reported rather than silently
changing another method's denominator.

```{r primary-metrics, echo=FALSE}
shown <- c("method", "heldout_qc_cv_mean", "residual_gam_deviance_mean",
           "residual_ljung_box_proportion", "batch_weighted_pc_r2",
           "biological_weighted_pc_r2_mean", "biological_associated_features",
           "cross_batch_effect_pearson_median", "total_runtime_sec",
           "features_retained")
knitr::kable(metrics[, shown], digits = 4, caption = "Primary retained-feature metrics")
```

## Denominator sensitivity

The common panel is the 111-feature intersection retained by all methods. The
coverage-penalized panel restores any feature dropped by a method to its Raw
value. Thus Raw held-out QC-CV is 32.60% on all 113 Raw features and 31.31% on
the 111-feature common panel; these are different estimands, not rerun drift.

```{r panels, echo=FALSE}
knitr::kable(metrics_common, digits = 4, caption = "Common 111-feature panel")
knitr::kable(metrics_coverage, digits = 4, caption = "Coverage-penalized 113-feature panel")
```

# WiNN selectivity and diagnostic equivalence

```{r winn, echo=FALSE}
knitr::kable(read_result("winn_selectivity_summary.csv"), digits = 4,
             caption = "Features and profiles actually modified by each WiNN mode")
knitr::kable(read_result("winn_instrumentation_validation.csv"), digits = 12,
             caption = "Instrumented-stage versus normal-wrapper matrix validation")
```

# Biological preservation

Primary age, BMI, and gender associations use limma on the common feature panel
after the same sample-wise osmolality division and log1p transform for every
method. Batch-adjusted models, batch-specific effect concordance, 100 bootstrap
resamples (seed 20260814), unnormalized sensitivity, and the prespecified
HU_neg_096_b2 exclusion are saved alongside the main table.

```{r biology, echo=FALSE}
assoc <- read_result("sacurine_associations_primary.csv")
assoc_summary <- aggregate(significant ~ method + variable, assoc, sum)
knitr::kable(assoc_summary, caption = "FDR-significant primary associations")
knitr::kable(read_result("sacurine_cross_batch_effect_concordance.csv"), digits = 4,
             caption = "Batch-pair effect concordance")
```

# Figures

```{r figures, echo=FALSE, out.width='96%'}
figure_names <- c(
  "acquisition_design.pdf", "randomization_diagnostics.pdf",
  "raw_vs_winn_pca.pdf", "technical_comparison.pdf",
  "biological_comparison.pdf", "winn_selectivity.pdf",
  "representative_feature_trajectories.pdf", "sacurine_effect_concordance.pdf"
)
paths <- file.path(result_dir, "figures", figure_names)
paths <- paths[file.exists(paths)]
knitr::include_graphics(paths, rel_path = FALSE)
```

# Reproducibility and limitations

The deposited table has 184 unique human study samples rather than biological
replicates across batches, so sample-replicate correlation and cross-batch
sample ICC are deliberately `NA`. Gender has a moderate within-batch order
association in batch ne1; biological findings should therefore be interpreted
jointly with the batch-adjusted and cross-batch sensitivity results. The
canonical qcrlscR 0.1.3 wrapper ignores its nominal span, so the benchmark uses
the separately audited, ID-safe per-batch primitive documented in
`reports/sacurine_qc_rlsc_audit.md`.

```{r session, echo=FALSE, results='asis'}
cat("```\n")
cat(paste(readLines(file.path(result_dir, "sessionInfo.txt"), warn = FALSE), collapse = "\n"))
cat("\n```\n")
```
