---
title: "Simulation Benchmark: Technical Correction Comparison"
author: "Tanmay Tanna <ttanna@ethz.ch>"
output:
  rmarkdown::html_notebook:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 180
)

options(
  width = 130,
  dplyr.summarise.inform = FALSE,
  pdf.useDingbats = FALSE
)

resolve_repo_root <- function() {
  candidates <- c(".", "..")
  for (cand in candidates) {
    if (
      file.exists(file.path(cand, "LICENSE")) &&
      dir.exists(file.path(cand, "data")) &&
      dir.exists(file.path(cand, "notebooks"))
    ) {
      return(normalizePath(cand, mustWork = TRUE))
    }
  }
  stop("Could not locate repository root.")
}

repo_root <- resolve_repo_root()
```

This notebook benchmarks technical correction methods on the simulated dataset in `data/simulated/`. It runs the full-method comparison and reports held-out QC-CV, residual run-order GAM deviance, per-feature ICC(A,1), per-sample Pearson correlation, and post-correction Ljung-Box autocorrelation counts.

## 1. Dependencies

```{r dependencies}
ensure_cran_package <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to load required CRAN package: ", pkg)
  }
}

ensure_bioc_package <- function(pkg) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to load required Bioconductor package: ", pkg)
  }
}

ensure_github_package <- function(pkg, repo) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    ensure_cran_package("remotes")
    remotes::install_github(repo, upgrade = "never")
  }
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Unable to install required GitHub package: ", pkg, " from ", repo)
  }
}

cran_pkgs <- c(
  "ggplot2", "dplyr", "tidyr", "purrr", "scales", "glue", "tibble",
  "readr", "qcrlscR", "pmartR", "rlang", "mgcv"
)
invisible(lapply(cran_pkgs, ensure_cran_package))

ensure_bioc_package("sva")
ensure_bioc_package("statTarget")
ensure_bioc_package("limma")
ensure_github_package("malbacR", "pmartR/malbacR")
ensure_github_package("TIGERr", "HAN-Siyu/TIGER")
ensure_github_package("winn", "ratschlab/winn")

library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(scales)
library(glue)
library(tibble)
library(readr)
library(sva)
library(statTarget)
library(qcrlscR)
library(pmartR)
library(malbacR)
library(TIGERr)
library(limma)
library(winn)
```

## 2. Styling And Global Objects

```{r style}
plot_dir <- file.path(repo_root, "notebooks", ".cache", "simulation_comparison")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

method_order_all <- c(
  "Raw",
  "ComBat",
  "QC-RLSC",
  "QC-RFSC",
  "TIGER",
  "SERRF",
  "WINN auto (QC)",
  "WINN auto-batch (QC)",
  "WINN default (no QC)"
)

method_palette <- c(
  "Raw" = "#4D4D4D",
  "ComBat" = "#0072B2",
  "QC-RLSC" = "#009E73",
  "QC-RFSC" = "#E69F00",
  "TIGER" = "#D55E00",
  "SERRF" = "#CC79A7",
  "WINN auto (QC)" = "#56B4E9",
  "WINN auto-batch (QC)" = "#F0E442",
  "WINN default (no QC)" = "#000000"
)

format_value_label <- function(x, digits = 2, integer_threshold = 1e-10) {
  ifelse(abs(x - round(x)) < integer_threshold, sprintf("%.0f", round(x)), sprintf(paste0("%.", digits, "f"), x))
}

format_mean_se_label <- function(mean_x, se_x, digits = 3) {
  ifelse(
    is.finite(mean_x) & is.finite(se_x),
    sprintf(paste0("%.", digits, "f \u00b1 %.", digits, "f"), mean_x, se_x),
    NA_character_
  )
}

build_contiguous_run_segments <- function(meta_df, sample_id_col = "sample_id", order_col = "run_order", batch_col = "batch") {
  md <- meta_df[order(meta_df[[order_col]], na.last = TRUE), , drop = FALSE]
  if (nrow(md) == 0L) {
    md$segment_index <- integer(0)
    md$segment_id <- character(0)
    return(md)
  }

  batch_vals <- as.character(md[[batch_col]])
  segment_index <- cumsum(c(TRUE, batch_vals[-1] != batch_vals[-length(batch_vals)]))

  md$segment_index <- segment_index
  md$segment_id <- paste0("batch_", batch_vals, "_segment_", sprintf("%03d", segment_index))
  md
}

fit_run_order_gam_explained <- function(series, run_order, min_obs = 6L, k_max = 6L) {
  ok <- is.finite(series) & is.finite(run_order)
  n_obs <- sum(ok)

  if (n_obs < min_obs) {
    return(c(explained = NA_real_, n_obs = n_obs, k = NA_real_))
  }

  y <- as.numeric(series[ok])
  x <- as.numeric(run_order[ok])
  n_unique_x <- length(unique(x))

  if (n_unique_x < 4L) {
    return(c(explained = NA_real_, n_obs = n_obs, k = NA_real_))
  }

  if (stats::sd(y) <= .Machine$double.eps) {
    return(c(explained = 0, n_obs = n_obs, k = NA_real_))
  }

  k_use <- min(as.integer(k_max), n_unique_x - 1L, max(3L, floor(n_obs / 2L)))
  if (!is.finite(k_use) || k_use < 3L) {
    return(c(explained = NA_real_, n_obs = n_obs, k = NA_real_))
  }

  fit <- tryCatch(
    mgcv::gam(y ~ s(x, bs = "cr", k = k_use), method = "REML"),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(c(explained = NA_real_, n_obs = n_obs, k = as.numeric(k_use)))
  }

  fit_summary <- tryCatch(summary(fit), error = function(e) NULL)
  if (is.null(fit_summary)) {
    return(c(explained = NA_real_, n_obs = n_obs, k = as.numeric(k_use)))
  }

  explained <- if (is.finite(fit_summary$dev.expl)) fit_summary$dev.expl else fit_summary$r.sq
  explained <- if (is.finite(explained)) min(max(as.numeric(explained), 0), 1) else NA_real_

  c(explained = explained, n_obs = n_obs, k = as.numeric(k_use))
}

compute_metabolite_segment_gam <- function(
  mat,
  meta_df,
  method_label,
  sample_id_col = "sample_id",
  order_col = "run_order",
  batch_col = "batch",
  transform_fun = log1p,
  min_obs = 6L,
  k_max = 6L
) {
  md <- meta_df[meta_df[[sample_id_col]] %in% colnames(mat), , drop = FALSE]
  md <- build_contiguous_run_segments(md, sample_id_col = sample_id_col, order_col = order_col, batch_col = batch_col)

  if (nrow(md) == 0L || nrow(mat) == 0L) {
    return(tibble::tibble(
      method = character(),
      metabolite = character(),
      batch = character(),
      segment_id = character(),
      n_obs = integer(),
      k = numeric(),
      explained = numeric()
    ))
  }

  seg_ids <- unique(md$segment_id)
  transformed <- transform_fun(mat[, md[[sample_id_col]], drop = FALSE])

  dplyr::bind_rows(lapply(seg_ids, function(seg_id) {
    seg_md <- md[md$segment_id == seg_id, , drop = FALSE]
    seg_sample_ids <- seg_md[[sample_id_col]]
    seg_mat <- transformed[, seg_sample_ids, drop = FALSE]

    gam_stats <- t(vapply(
      seq_len(nrow(seg_mat)),
      function(i) fit_run_order_gam_explained(seg_mat[i, ], seg_md[[order_col]], min_obs = min_obs, k_max = k_max),
      numeric(3)
    ))

    tibble::tibble(
      method = method_label,
      metabolite = rownames(seg_mat),
      batch = as.character(seg_md[[batch_col]][1]),
      segment_id = seg_id,
      n_obs = as.integer(gam_stats[, "n_obs"]),
      k = as.numeric(gam_stats[, "k"]),
      explained = as.numeric(gam_stats[, "explained"])
    )
  }))
}

summarise_metabolite_segment_gam <- function(df, digits = 4) {
  df |>
    dplyr::group_by(method) |>
    dplyr::summarise(
      n_pairs = sum(is.finite(explained)),
      mean_explained = mean(explained, na.rm = TRUE),
      median_explained = stats::median(explained, na.rm = TRUE),
      sd_explained = stats::sd(explained, na.rm = TRUE),
      mean_sd = ifelse(
        is.finite(mean_explained),
        sprintf(paste0("%.", digits, "f \u00b1 %.", digits, "f"), mean_explained, ifelse(is.finite(sd_explained), sd_explained, NA_real_)),
        NA_character_
      ),
      .groups = "drop"
    ) |>
    arrange(method)
}

ljung_box_series_metrics <- function(series, max_lag = 10L, min_obs = 8L) {
  x <- as.numeric(series)
  x <- x[is.finite(x)]
  n_obs <- length(x)
  lag_use <- min(as.integer(max_lag), max(1L, floor(n_obs / 3L)), n_obs - 1L)

  if (n_obs < min_obs || lag_use < 1L || isTRUE(stats::sd(x, na.rm = TRUE) < .Machine$double.eps)) {
    return(c(lb_stat = NA_real_, p_value = NA_real_, lag = as.numeric(lag_use), n_obs = as.numeric(n_obs)))
  }

  lb <- tryCatch(
    stats::Box.test(x, lag = lag_use, type = "Ljung-Box"),
    error = function(e) NULL
  )

  if (is.null(lb)) {
    return(c(lb_stat = NA_real_, p_value = NA_real_, lag = as.numeric(lag_use), n_obs = as.numeric(n_obs)))
  }

  c(
    lb_stat = as.numeric(lb$statistic),
    p_value = as.numeric(lb$p.value),
    lag = as.numeric(lag_use),
    n_obs = as.numeric(n_obs)
  )
}

compute_metabolite_segment_ljung_box <- function(
  mat,
  meta_df,
  method_label,
  sample_id_col = "sample_id",
  order_col = "run_order",
  batch_col = "batch",
  transform_fun = log1p,
  min_obs = 8L,
  max_lag = 10L
) {
  md <- meta_df[meta_df[[sample_id_col]] %in% colnames(mat), , drop = FALSE]
  md <- build_contiguous_run_segments(md, sample_id_col = sample_id_col, order_col = order_col, batch_col = batch_col)

  if (nrow(md) == 0L || nrow(mat) == 0L) {
    return(tibble::tibble(
      method = character(),
      metabolite = character(),
      batch = character(),
      segment_id = character(),
      n_obs = integer(),
      lag = numeric(),
      lb_stat = numeric(),
      p_value = numeric()
    ))
  }

  seg_ids <- unique(md$segment_id)
  transformed <- transform_fun(mat[, md[[sample_id_col]], drop = FALSE])

  dplyr::bind_rows(lapply(seg_ids, function(seg_id) {
    seg_md <- md[md$segment_id == seg_id, , drop = FALSE]
    seg_sample_ids <- seg_md[[sample_id_col]]
    seg_mat <- transformed[, seg_sample_ids, drop = FALSE]

    lb_stats <- t(vapply(
      seq_len(nrow(seg_mat)),
      function(i) ljung_box_series_metrics(seg_mat[i, ], max_lag = max_lag, min_obs = min_obs),
      numeric(4)
    ))

    tibble::tibble(
      method = method_label,
      metabolite = rownames(seg_mat),
      batch = as.character(seg_md[[batch_col]][1]),
      segment_id = seg_id,
      n_obs = as.integer(lb_stats[, "n_obs"]),
      lag = as.numeric(lb_stats[, "lag"]),
      lb_stat = as.numeric(lb_stats[, "lb_stat"]),
      p_value = as.numeric(lb_stats[, "p_value"])
    )
  }))
}

annotate_metabolite_segment_ljung_box <- function(df, alpha = 0.05, adjust_method = "BH") {
  df |>
    dplyr::group_by(method) |>
    dplyr::mutate(
      p_adj = stats::p.adjust(p_value, method = adjust_method),
      is_autocorrelated = is.finite(p_adj) & (p_adj < alpha)
    ) |>
    dplyr::ungroup()
}

summarise_metabolite_segment_ljung_box <- function(df) {
  df |>
    dplyr::group_by(method) |>
    dplyr::summarise(
      n_tested = sum(is.finite(p_value)),
      n_autocorrelated = sum(is_autocorrelated, na.rm = TRUE),
      prop_autocorrelated = ifelse(n_tested > 0, n_autocorrelated / n_tested, NA_real_),
      median_lb_stat = stats::median(lb_stat, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(method)
}
```

## 3. Data Load And Validation

```{r data}
simulated_dir <- file.path(repo_root, "data", "simulated")
gt_path <- file.path(simulated_dir, "ground_truth.tsv")
shift_path <- file.path(simulated_dir, "shifted_profiles.tsv")
meta_path <- file.path(simulated_dir, "sample_metadata.tsv")

if (!file.exists(gt_path)) stop("Missing simulated data file: ", gt_path)
if (!file.exists(shift_path)) stop("Missing simulated data file: ", shift_path)
if (!file.exists(meta_path)) stop("Missing simulated data file: ", meta_path)

gt_tbl <- readr::read_tsv(gt_path, show_col_types = FALSE)
shift_tbl <- readr::read_tsv(shift_path, show_col_types = FALSE)
meta <- readr::read_tsv(meta_path, show_col_types = FALSE)

tbl_to_mat <- function(tbl) {
  feat_col <- colnames(tbl)[1]
  m <- as.matrix(tbl[, -1, drop = FALSE])
  storage.mode(m) <- "double"
  rownames(m) <- tbl[[feat_col]]
  colnames(m) <- colnames(tbl)[-1]
  m
}

GT <- tbl_to_mat(gt_tbl)
x <- tbl_to_mat(shift_tbl)

meta <- meta |>
  mutate(
    sample_id = as.character(sample_id),
    plate = as.character(plate),
    batch = as.integer(sub("^P", "", plate)),
    class = if_else(sample_type == "control", "QC", "Sample")
  ) |>
  arrange(run_order)

common_samples <- Reduce(intersect, list(meta$sample_id, colnames(GT), colnames(x)))
if (length(common_samples) < 10L) {
  stop("Insufficient shared sample IDs between metadata and matrices.")
}

common_features <- intersect(rownames(GT), rownames(x))
if (length(common_features) < 10L) {
  stop("Insufficient shared metabolites between GT and shifted matrix.")
}

meta <- meta[match(common_samples, meta$sample_id), , drop = FALSE]
GT <- GT[common_features, common_samples, drop = FALSE]
x <- x[common_features, common_samples, drop = FALSE]

qc_ids <- meta$sample_id[meta$class == "QC"]
non_qc_ids <- meta$sample_id[meta$class != "QC"]

if (length(qc_ids) < 6L) stop("Need at least 6 QC samples for QC-CV evaluation.")
if (length(non_qc_ids) < 10L) stop("Need at least 10 non-QC samples for correlation evaluation.")

cohort_summary <- tibble(
  Metric = c("Total samples", "QC samples", "Non-QC samples", "Metabolites", "Batches"),
  Value = c(ncol(x), length(qc_ids), length(non_qc_ids), nrow(x), dplyr::n_distinct(meta$batch))
)

knitr::kable(cohort_summary, align = c("l", "r"))
```

## 4. Method Helpers

```{r method-helpers}
clip_nonnegative <- function(m) {
  m <- as.matrix(m)
  storage.mode(m) <- "double"
  m[is.infinite(m)] <- NA_real_
  pmax(m, 0)
}

run_silently <- function(code) {
  suppressWarnings(suppressMessages(force(code)))
}

safe_cor <- function(a, b, method = "pearson", min_n = 6L) {
  ok <- is.finite(a) & is.finite(b)
  if (sum(ok) < min_n) return(NA_real_)
  if (stats::sd(a[ok]) <= 0 || stats::sd(b[ok]) <= 0) return(NA_real_)
  suppressWarnings(cor(a[ok], b[ok], method = method))
}

calc_icc_a1_matrix <- function(ratings) {
  ratings <- as.matrix(ratings)
  ratings <- ratings[complete.cases(ratings), , drop = FALSE]

  n <- nrow(ratings)
  k <- ncol(ratings)
  if (n < 2L || k < 2L) {
    return(NA_real_)
  }

  grand_mean <- mean(ratings)
  row_means <- rowMeans(ratings)
  col_means <- colMeans(ratings)

  ss_rows <- k * sum((row_means - grand_mean)^2)
  ss_cols <- n * sum((col_means - grand_mean)^2)
  ss_total <- sum((ratings - grand_mean)^2)
  ss_error <- ss_total - ss_rows - ss_cols

  ms_rows <- ss_rows / (n - 1L)
  ms_cols <- ss_cols / (k - 1L)
  ms_error <- ss_error / ((n - 1L) * (k - 1L))

  denom <- ms_rows + (k - 1L) * ms_error + (k * (ms_cols - ms_error) / n)
  if (!is.finite(denom) || denom <= 0) {
    return(NA_real_)
  }

  (ms_rows - ms_error) / denom
}

calc_icc_a1_pair <- function(reference, candidate, min_n = 6L) {
  dat <- cbind(reference, candidate)
  dat <- dat[complete.cases(dat), , drop = FALSE]
  if (nrow(dat) < min_n) {
    return(NA_real_)
  }
  calc_icc_a1_matrix(dat)
}

align_meta_to_mat <- function(x_mat, meta_df = meta) {
  sample_ids <- colnames(x_mat)
  md <- meta_df[match(sample_ids, meta_df$sample_id), , drop = FALSE]
  if (anyNA(md$sample_id)) {
    stop("Sample IDs in matrix do not match metadata sample IDs.")
  }
  md
}

attach_qc_passthrough <- function(corrected_mat, raw_mat, meta_df) {
  qc_samples <- meta_df$sample_id[meta_df$class == "QC" & meta_df$sample_id %in% colnames(raw_mat)]
  if (length(qc_samples) == 0L) {
    return(corrected_mat)
  }

  feature_ids <- intersect(rownames(corrected_mat), rownames(raw_mat))
  corrected_mat <- corrected_mat[feature_ids, , drop = FALSE]
  raw_aligned <- raw_mat[feature_ids, , drop = FALSE]

  missing_qc <- setdiff(qc_samples, colnames(corrected_mat))
  if (length(missing_qc) > 0L) {
    corrected_mat <- cbind(corrected_mat, raw_aligned[, missing_qc, drop = FALSE])
  }

  output_ids <- meta_df$sample_id[meta_df$sample_id %in% colnames(corrected_mat)]
  corrected_mat[, output_ids, drop = FALSE]
}

run_method <- function(label, fn) {
  elapsed <- system.time({
    out <- fn()
  })[["elapsed"]]

  list(
    label = label,
    data = clip_nonnegative(out),
    elapsed_sec = elapsed
  )
}

run_method_cached <- function(label, fn, force = FALSE, cache_tag = "") {
  cache_dir <- file.path(plot_dir, "method_cache")
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

  safe_label <- gsub("[^A-Za-z0-9]+", "_", tolower(label))
  tag <- ifelse(nzchar(cache_tag), paste0("_", cache_tag), "")
  cache_file <- file.path(cache_dir, paste0(safe_label, tag, ".rds"))

  if (!isTRUE(force) && file.exists(cache_file)) {
    return(readRDS(cache_file))
  }

  out <- run_method(label = label, fn = fn)
  saveRDS(out, cache_file)
  out
}

run_matrix_cached <- function(label, fn, force = FALSE, cache_tag = "") {
  cache_dir <- file.path(plot_dir, "matrix_cache")
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

  safe_label <- gsub("[^A-Za-z0-9]+", "_", tolower(label))
  tag <- ifelse(nzchar(cache_tag), paste0("_", cache_tag), "")
  cache_file <- file.path(cache_dir, paste0(safe_label, tag, ".rds"))

  if (!isTRUE(force) && file.exists(cache_file)) {
    return(readRDS(cache_file))
  }

  out <- clip_nonnegative(fn())
  saveRDS(out, cache_file)
  out
}

with_no_cluster <- function(code, patch_do_parallel = FALSE) {
  parallel_ns <- asNamespace("parallel")
  original_parallel <- list(
    makeCluster = get("makeCluster", envir = parallel_ns),
    clusterExport = get("clusterExport", envir = parallel_ns),
    stopCluster = get("stopCluster", envir = parallel_ns)
  )

  if (bindingIsLocked("makeCluster", parallel_ns)) unlockBinding("makeCluster", parallel_ns)
  if (bindingIsLocked("clusterExport", parallel_ns)) unlockBinding("clusterExport", parallel_ns)
  if (bindingIsLocked("stopCluster", parallel_ns)) unlockBinding("stopCluster", parallel_ns)

  assign("makeCluster", function(spec, ...) NULL, envir = parallel_ns)
  assign("clusterExport", function(cl, ...) invisible(NULL), envir = parallel_ns)
  assign("stopCluster", function(cl, ...) invisible(NULL), envir = parallel_ns)

  lockBinding("makeCluster", parallel_ns)
  lockBinding("clusterExport", parallel_ns)
  lockBinding("stopCluster", parallel_ns)

  do_parallel_ns <- NULL
  original_register <- NULL

  if (isTRUE(patch_do_parallel) && requireNamespace("doParallel", quietly = TRUE)) {
    do_parallel_ns <- asNamespace("doParallel")
    original_register <- get("registerDoParallel", envir = do_parallel_ns)
    if (bindingIsLocked("registerDoParallel", do_parallel_ns)) unlockBinding("registerDoParallel", do_parallel_ns)
    assign("registerDoParallel", function(cl, ...) foreach::registerDoSEQ(), envir = do_parallel_ns)
    lockBinding("registerDoParallel", do_parallel_ns)
  }

  on.exit({
    if (!is.null(do_parallel_ns) && !is.null(original_register)) {
      if (bindingIsLocked("registerDoParallel", do_parallel_ns)) unlockBinding("registerDoParallel", do_parallel_ns)
      assign("registerDoParallel", original_register, envir = do_parallel_ns)
      lockBinding("registerDoParallel", do_parallel_ns)
    }

    if (bindingIsLocked("makeCluster", parallel_ns)) unlockBinding("makeCluster", parallel_ns)
    if (bindingIsLocked("clusterExport", parallel_ns)) unlockBinding("clusterExport", parallel_ns)
    if (bindingIsLocked("stopCluster", parallel_ns)) unlockBinding("stopCluster", parallel_ns)

    assign("makeCluster", original_parallel$makeCluster, envir = parallel_ns)
    assign("clusterExport", original_parallel$clusterExport, envir = parallel_ns)
    assign("stopCluster", original_parallel$stopCluster, envir = parallel_ns)

    lockBinding("makeCluster", parallel_ns)
    lockBinding("clusterExport", parallel_ns)
    lockBinding("stopCluster", parallel_ns)
  }, add = TRUE)

  force(code)
}

make_malbac_object <- function(x_mat, meta_df, jitter_eps = 0) {
  x_mat <- as.matrix(x_mat)
  if (jitter_eps > 0) {
    x_mat <- x_mat + matrix(runif(length(x_mat), min = 0, max = jitter_eps), nrow = nrow(x_mat))
  }

  safe_sample <- paste0("Sample_", seq_len(ncol(x_mat)))
  sample_map <- setNames(meta_df$sample_id, safe_sample)

  x_safe <- x_mat
  colnames(x_safe) <- safe_sample

  f_data <- data.frame(
    sampleID = safe_sample,
    sample_type = meta_df$class,
    batch = meta_df$batch,
    run_order = meta_df$run_order,
    stringsAsFactors = FALSE
  )
  rownames(f_data) <- f_data$sampleID

  e_data <- data.frame(
    feature = rownames(x_safe),
    x_safe,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

  obj <- pmartR::as.metabData(
    e_data = e_data,
    f_data = f_data,
    edata_cname = "feature",
    fdata_cname = "sampleID"
  )

  obj <- pmartR::group_designation(
    omicsData = obj,
    main_effects = "sample_type",
    batch_id = "batch"
  )

  list(obj = obj, sample_map = sample_map)
}

extract_malbac_matrix <- function(malbac_obj, sample_map) {
  e_data <- malbac_obj$e_data
  m <- as.matrix(e_data[, -1, drop = FALSE])
  rownames(m) <- e_data[, 1]

  mapped <- sample_map[colnames(m)]
  colnames(m) <- unname(mapped)
  m
}

run_combat <- function(x_mat, meta_df = meta, par_prior = TRUE) {
  md <- align_meta_to_mat(x_mat, meta_df)
  x_log <- log1p(x_mat)
  y_log <- sva::ComBat(
    dat = x_log,
    batch = md$batch,
    mod = NULL,
    par.prior = par_prior,
    prior.plots = FALSE
  )
  expm1(y_log)
}

run_qc_rlsc_with_controls <- function(x_mat, control_ids, meta_df = meta, span = 0.75) {
  md <- align_meta_to_mat(x_mat, meta_df) |>
    arrange(run_order)
  x_ord <- x_mat[, md$sample_id, drop = FALSE]
  dat <- as.data.frame(t(x_ord), check.names = FALSE)
  y <- ifelse(md$sample_id %in% control_ids, "QC", "Sample")
  corrected <- qcrlscR::qc.rlsc.wrap(
    dat = dat,
    cls.qc = y,
    cls.bl = factor(md$batch),
    method = "divide",
    intra = TRUE,
    opti = FALSE,
    log10 = FALSE,
    outl = FALSE,
    shift = TRUE,
    span = span
  )
  corrected <- t(as.matrix(corrected))
  rownames(corrected) <- rownames(x_ord)
  colnames(corrected) <- md$sample_id
  corrected[, meta_df$sample_id, drop = FALSE]
}

run_qc_rfsc_with_controls <- function(x_mat, control_ids, meta_df = meta, ntree = 500, coCV = 30, Frule = 0.8) {
  md <- align_meta_to_mat(x_mat, meta_df) |>
    arrange(run_order)
  x_ord <- x_mat[, md$sample_id, drop = FALSE]

  control_ids <- intersect(control_ids, md$sample_id)
  if (length(control_ids) < 2L) {
    stop("QC-RFSC requires at least 2 QC controls.")
  }

  first_id <- md$sample_id[1]
  last_id <- md$sample_id[nrow(md)]
  if (!(first_id %in% control_ids) || !(last_id %in% control_ids)) {
    first_qc <- control_ids[which.min(md$run_order[match(control_ids, md$sample_id)])]
    last_qc <- control_ids[which.max(md$run_order[match(control_ids, md$sample_id)])]
    reordered_ids <- c(first_qc, setdiff(md$sample_id, c(first_qc, last_qc)), last_qc)
    md <- md[match(reordered_ids, md$sample_id), , drop = FALSE]
    x_ord <- x_ord[, reordered_ids, drop = FALSE]
  }

  qc_flag <- md$sample_id %in% control_ids
  stat_names <- ifelse(qc_flag, paste0(md$sample_id, "_QC"), md$sample_id)

  samPeno <- data.frame(
    sample = stat_names,
    batch = md$batch,
    class = ifelse(qc_flag, NA_integer_, 1L),
    order = seq_len(nrow(md)),
    stringsAsFactors = FALSE
  )

  if (!is.na(samPeno$class[1]) || !is.na(samPeno$class[nrow(samPeno)])) {
    stop("QC-RFSC setup failed to place QC at first and last positions.")
  }

  samFile <- data.frame(name = rownames(x_ord), x_ord, check.names = FALSE)
  colnames(samFile)[-1] <- stat_names

  tmp <- tempfile("sim_shiftcor_")
  dir.create(tmp)
  old <- getwd()
  setwd(tmp)
  on.exit(setwd(old), add = TRUE)

  write.csv(samPeno, "samPeno.csv", row.names = FALSE, quote = FALSE)
  write.csv(samFile, "samFile.csv", row.names = FALSE, quote = FALSE)

  run_silently(
    statTarget::shiftCor(
      samPeno = "samPeno.csv",
      samFile = "samFile.csv",
      Frule = Frule,
      MLmethod = "QCRFSC",
      ntree = ntree,
      coCV = coCV,
      plot = FALSE
    )
  )

  out_file <- file.path("statTarget", "shiftCor", "After_shiftCor", "shift_all_cor.csv")
  if (!file.exists(out_file)) stop("statTarget did not produce shift_all_cor.csv")

  corrected_df <- read.csv(out_file, check.names = FALSE, row.names = 1, stringsAsFactors = FALSE)
  if ("class" %in% rownames(corrected_df)) {
    corrected_df <- corrected_df[rownames(corrected_df) != "class", , drop = FALSE]
  }

  corrected <- apply(corrected_df, 2, as.numeric)
  rownames(corrected) <- rownames(corrected_df)
  colnames(corrected) <- sub("_QC$", "", colnames(corrected_df))
  corrected <- corrected / 1000

  corrected[, colnames(x_mat), drop = FALSE]
}

run_tiger <- function(x_mat, meta_df = meta, use_injection = TRUE, include_qc_passthrough = TRUE) {
  md <- align_meta_to_mat(x_mat, meta_df)

  sample_df <- tibble(
    sampleID = md$sample_id,
    sampleType = md$class,
    batchID = md$batch,
    injectionOrder = md$run_order
  ) |>
    bind_cols(as.data.frame(t(x_mat), check.names = FALSE))

  train_samples <- sample_df |> filter(sampleType == "QC")
  if (nrow(train_samples) < 2L) {
    stop("TIGER requires at least two QC samples in training data.")
  }

  test_samples <- sample_df |> filter(sampleType != "QC")

  if (!isTRUE(use_injection)) {
    train_samples <- train_samples |> select(-injectionOrder)
    test_samples <- test_samples |> select(-injectionOrder)
  }

  corrected <- run_silently(
    TIGERr::run_TIGER(
      test_samples = test_samples,
      train_samples = train_samples,
      col_sampleID = "sampleID",
      col_sampleType = "sampleType",
      col_batchID = "batchID",
      col_order = if (isTRUE(use_injection)) "injectionOrder" else NULL
    )
  )

  meta_cols <- c("sampleID", "sampleType", "batchID")
  if (isTRUE(use_injection)) {
    meta_cols <- c(meta_cols, "injectionOrder")
  }

  feature_cols <- setdiff(colnames(corrected), meta_cols)
  m <- t(as.matrix(corrected[, feature_cols, drop = FALSE]))
  rownames(m) <- feature_cols
  colnames(m) <- corrected$sampleID

  if (isTRUE(include_qc_passthrough)) {
    attach_qc_passthrough(m, x_mat, md)
  } else {
    m
  }
}

run_tiger_with_holdout_qc <- function(
  x_mat,
  train_qc_ids,
  test_qc_ids,
  meta_df = meta,
  use_injection = TRUE,
  mtry_percent = 0.4,
  nodesize_percent = 0.4,
  ntree = 5,
  parallel_cores = 2,
  duplicate_test_qc = TRUE,
  include_nonqc_test = TRUE
) {
  md <- align_meta_to_mat(x_mat, meta_df)

  sample_df <- tibble(
    sampleID = md$sample_id,
    sampleType = md$class,
    batchID = md$batch,
    injectionOrder = md$run_order
  ) |>
    bind_cols(as.data.frame(t(x_mat), check.names = FALSE))

  train_qc_ids <- intersect(train_qc_ids, md$sample_id[md$class == "QC"])
  test_qc_ids <- intersect(test_qc_ids, md$sample_id[md$class == "QC"])
  if (length(train_qc_ids) < 2L) stop("TIGER holdout run requires at least 2 training QCs.")
  if (length(test_qc_ids) < 1L) stop("TIGER holdout run requires at least 1 test QC.")

  train_samples <- sample_df |> filter(sampleID %in% train_qc_ids)
  if (isTRUE(duplicate_test_qc)) {
    nonqc_test <- if (isTRUE(include_nonqc_test)) {
      sample_df |> filter(sampleType != "QC")
    } else {
      sample_df[0, , drop = FALSE]
    }
    qc_test_dup <- sample_df |>
      filter(sampleID %in% test_qc_ids) |>
      mutate(sampleID = paste0("QCtest__", sampleID), sampleType = "QC_test")
    test_samples <- bind_rows(nonqc_test, qc_test_dup)
  } else {
    test_ids <- if (isTRUE(include_nonqc_test)) {
      unique(c(md$sample_id[md$class != "QC"], test_qc_ids))
    } else {
      unique(test_qc_ids)
    }
    test_samples <- sample_df |> filter(sampleID %in% test_ids)
  }

  if (!isTRUE(use_injection)) {
    train_samples <- train_samples |> select(-injectionOrder)
    test_samples <- test_samples |> select(-injectionOrder)
  }

  corrected <- run_silently(
    TIGERr::run_TIGER(
      test_samples = test_samples,
      train_samples = train_samples,
      col_sampleID = "sampleID",
      col_sampleType = "sampleType",
      col_batchID = "batchID",
      col_order = if (isTRUE(use_injection)) "injectionOrder" else NULL,
      mtry_percent = mtry_percent,
      nodesize_percent = nodesize_percent,
      ntree = ntree,
      parallel.cores = parallel_cores
    )
  )

  meta_cols <- c("sampleID", "sampleType", "batchID")
  if (isTRUE(use_injection)) {
    meta_cols <- c(meta_cols, "injectionOrder")
  }

  feature_cols <- setdiff(colnames(corrected), meta_cols)
  m <- t(as.matrix(corrected[, feature_cols, drop = FALSE]))
  rownames(m) <- feature_cols
  colnames(m) <- corrected$sampleID

  if (isTRUE(duplicate_test_qc)) {
    dup_cols <- grep("^QCtest__", colnames(m), value = TRUE)
    if (length(dup_cols) > 0L) {
      m_qc <- m[, dup_cols, drop = FALSE]
      colnames(m_qc) <- sub("^QCtest__", "", dup_cols)
      nondup_cols <- setdiff(colnames(m), dup_cols)
      m <- cbind(m[, nondup_cols, drop = FALSE], m_qc)
    }
  }

  m
}

run_serrf <- function(x_mat, meta_df = meta, jitter_eps = 0, include_qc_passthrough = TRUE) {
  md <- align_meta_to_mat(x_mat, meta_df)

  jitter_candidates <- unique(c(
    as.numeric(jitter_eps),
    if (isTRUE(jitter_eps <= 0)) c(1e-6, 1e-4) else c(as.numeric(jitter_eps) * 10)
  ))

  out <- NULL
  sample_map <- NULL
  last_err <- NULL
  for (jit in jitter_candidates) {
    mal <- make_malbac_object(x_mat, md, jitter_eps = jit)
    out_try <- tryCatch(
      run_silently(
        with_no_cluster(
          malbacR::bc_serrf(
            omicsData = mal$obj,
            sampletype_cname = "sample_type",
            test_val = "QC",
            group_cname = "sample_type"
          ),
          patch_do_parallel = TRUE
        )
      ),
      error = function(e) e
    )

    if (!inherits(out_try, "error")) {
      out <- out_try
      sample_map <- mal$sample_map
      break
    }
    last_err <- out_try
  }

  if (is.null(out)) {
    stop(last_err)
  }

  corrected <- extract_malbac_matrix(out, sample_map)
  corrected <- corrected[, md$sample_id[md$sample_id %in% colnames(corrected)], drop = FALSE]

  if (!isTRUE(include_qc_passthrough)) {
    return(corrected)
  }

  attach_qc_passthrough(corrected, x_mat, md)
}

run_serrf_with_holdout_qc <- function(
  x_mat,
  train_qc_ids,
  test_qc_ids,
  meta_df = meta,
  jitter_eps = 0,
  duplicate_test_qc = TRUE,
  include_nonqc_samples = TRUE
) {
  md <- align_meta_to_mat(x_mat, meta_df)

  if (!isTRUE(include_nonqc_samples)) {
    keep_ids <- md$sample_id[md$class == "QC"]
    md <- md[match(keep_ids, md$sample_id), , drop = FALSE]
    x_mat <- x_mat[, keep_ids, drop = FALSE]
  }

  train_qc_ids <- intersect(train_qc_ids, md$sample_id[md$class == "QC"])
  test_qc_ids <- intersect(test_qc_ids, md$sample_id[md$class == "QC"])
  if (length(train_qc_ids) < 2L) stop("SERRF holdout run requires at least 2 training QCs.")
  if (length(test_qc_ids) < 1L) stop("SERRF holdout run requires at least 1 test QC.")

  md_serrf <- md
  md_serrf$class <- as.character(md_serrf$class)
  md_serrf$class[md_serrf$class == "QC"] <- "QC_holdout"
  md_serrf$class[md_serrf$sample_id %in% train_qc_ids] <- "QC"

  x_serrf <- x_mat
  if (isTRUE(duplicate_test_qc)) {
    dup_ids <- paste0("QCtest__", test_qc_ids)
    dup_block <- x_mat[, test_qc_ids, drop = FALSE]
    colnames(dup_block) <- dup_ids
    x_serrf <- cbind(x_serrf, dup_block)

    dup_meta <- md[match(test_qc_ids, md$sample_id), , drop = FALSE]
    dup_meta$sample_id <- dup_ids
    dup_meta$class <- "QC_test"
    dup_meta$run_order <- dup_meta$run_order + seq_along(test_qc_ids) * 1e-4
    md_serrf <- bind_rows(md_serrf, dup_meta)
  } else {
    md_serrf$class[md_serrf$sample_id %in% test_qc_ids] <- "QC_test"
  }

  md_serrf <- md_serrf[match(colnames(x_serrf), md_serrf$sample_id), , drop = FALSE]

  jitter_candidates <- unique(c(
    as.numeric(jitter_eps),
    if (isTRUE(jitter_eps <= 0)) c(1e-6, 1e-4) else c(as.numeric(jitter_eps) * 10)
  ))

  out <- NULL
  sample_map <- NULL
  last_err <- NULL
  for (jit in jitter_candidates) {
    mal <- make_malbac_object(x_serrf, md_serrf, jitter_eps = jit)
    out_try <- tryCatch(
      run_silently(
        with_no_cluster(
          malbacR::bc_serrf(
            omicsData = mal$obj,
            sampletype_cname = "sample_type",
            test_val = "QC",
            group_cname = "sample_type"
          ),
          patch_do_parallel = TRUE
        )
      ),
      error = function(e) e
    )

    if (!inherits(out_try, "error")) {
      out <- out_try
      sample_map <- mal$sample_map
      break
    }
    last_err <- out_try
  }

  if (is.null(out)) stop(last_err)

  m <- extract_malbac_matrix(out, sample_map)
  if (isTRUE(duplicate_test_qc)) {
    dup_cols <- paste0("QCtest__", test_qc_ids)
    dup_cols <- dup_cols[dup_cols %in% colnames(m)]
    if (length(dup_cols) > 0L) {
      m_qc <- m[, dup_cols, drop = FALSE]
      colnames(m_qc) <- sub("^QCtest__", "", dup_cols)
      nondup_cols <- setdiff(colnames(m), dup_cols)
      m <- cbind(m[, nondup_cols, drop = FALSE], m_qc)
    }
  }

  m
}

run_winn_with_controls <- function(x_mat, control_ids = NULL, meta_df = meta, auto_batch = FALSE, parameters = "auto") {
  md <- align_meta_to_mat(x_mat, meta_df)
  control_idx_local <- if (is.null(control_ids)) NULL else which(md$sample_id %in% control_ids)

  winn::winn(
    data = x_mat,
    batch = if (isTRUE(auto_batch)) NULL else md$batch,
    run_order = md$run_order,
    control_samples = control_idx_local,
    parameters = parameters
  )
}

run_winn_mode1_auto_qc <- function(x_mat, meta_df = meta) {
  run_winn_with_controls(x_mat, control_ids = meta_df$sample_id[meta_df$class == "QC"], meta_df = meta_df, auto_batch = FALSE, parameters = "auto")
}

run_winn_mode2_auto_batch_qc <- function(x_mat, meta_df = meta) {
  run_winn_with_controls(x_mat, control_ids = meta_df$sample_id[meta_df$class == "QC"], meta_df = meta_df, auto_batch = TRUE, parameters = "auto")
}

run_winn_mode3_default_no_qc <- function(x_mat, meta_df = meta) {
  run_winn_with_controls(x_mat, control_ids = NULL, meta_df = meta_df, auto_batch = FALSE, parameters = "fixed")
}
```

## 5. Run Methods

```{r run-methods, cache=TRUE}
method_results <- list(
  run_method_cached("Raw", function() x),
  run_method_cached("ComBat", function() run_combat(x, meta_df = meta)),
  run_method_cached("QC-RLSC", function() run_qc_rlsc_with_controls(x, control_ids = qc_ids, meta_df = meta)),
  run_method_cached("QC-RFSC", function() run_qc_rfsc_with_controls(x, control_ids = qc_ids, meta_df = meta)),
  run_method_cached("TIGER", function() run_tiger(x, meta_df = meta, use_injection = TRUE, include_qc_passthrough = TRUE)),
  run_method_cached("SERRF", function() run_serrf(x, meta_df = meta, jitter_eps = 0, include_qc_passthrough = TRUE)),
  run_method_cached("WINN auto (QC)", function() run_winn_mode1_auto_qc(x, meta_df = meta)),
  run_method_cached("WINN auto-batch (QC)", function() run_winn_mode2_auto_batch_qc(x, meta_df = meta)),
  run_method_cached("WINN default (no QC)", function() run_winn_mode3_default_no_qc(x, meta_df = meta))
)

method_mats <- setNames(
  lapply(method_results, function(res) res$data),
  vapply(method_results, function(res) res$label, character(1))
)
```

## 6. Metrics

```{r metrics, cache=TRUE}
cv_percent <- function(mat, sample_ids) {
  sample_ids <- intersect(sample_ids, colnames(mat))
  if (length(sample_ids) < 3L || nrow(mat) < 2L) {
    return(setNames(rep(NA_real_, nrow(mat)), rownames(mat)))
  }

  sub <- mat[, sample_ids, drop = FALSE]
  mean_vec <- rowMeans(sub, na.rm = TRUE)
  sd_vec <- apply(sub, 1, sd, na.rm = TRUE)
  denom <- pmax(abs(mean_vec), .Machine$double.eps)
  out <- 100 * sd_vec / denom
  names(out) <- rownames(sub)
  out
}

set.seed(20260311)
holdout_qc_ids <- meta |>
  filter(class == "QC") |>
  group_by(batch) |>
  group_modify(~ {
    n_hold <- min(2L, max(0L, nrow(.x) - 3L))
    if (n_hold <= 0L) {
      return(.x[0, , drop = FALSE])
    }
    .x |>
      slice_sample(n = n_hold)
  }) |>
  ungroup() |>
  pull(sample_id)

train_qc_ids <- setdiff(qc_ids, holdout_qc_ids)
if (length(train_qc_ids) < 2L) stop("Insufficient train QCs after holdout split.")

holdout_tag_num <- abs(sum(utf8ToInt(paste(sort(holdout_qc_ids), collapse = "|"))))
holdout_tag <- paste0("holdout_", holdout_tag_num)

qccv_tiger_holdout <- run_matrix_cached(
  "TIGER_QCCV_HOLDOUT",
  function() run_tiger_with_holdout_qc(
    x_mat = x,
    train_qc_ids = train_qc_ids,
    test_qc_ids = holdout_qc_ids,
    meta_df = meta,
    use_injection = TRUE,
    mtry_percent = 0.4,
    nodesize_percent = 0.4,
    ntree = 5,
    parallel_cores = 2,
    duplicate_test_qc = TRUE,
    include_nonqc_test = TRUE
  ),
  cache_tag = paste0(holdout_tag, "_ntree5_nonqc1")
)

qccv_serrf_holdout <- run_matrix_cached(
  "SERRF_QCCV_HOLDOUT",
  function() run_serrf_with_holdout_qc(
    x_mat = x,
    train_qc_ids = train_qc_ids,
    test_qc_ids = holdout_qc_ids,
    meta_df = meta,
    jitter_eps = 0,
    duplicate_test_qc = TRUE,
    include_nonqc_samples = TRUE
  ),
  cache_tag = holdout_tag
)

qccv_holdout_mats <- list(
  "Raw" = method_mats[["Raw"]],
  "ComBat" = method_mats[["ComBat"]],
  "QC-RLSC" = run_matrix_cached(
    "QC-RLSC_HOLDOUT",
    function() run_qc_rlsc_with_controls(x, control_ids = train_qc_ids, meta_df = meta),
    cache_tag = holdout_tag
  ),
  "QC-RFSC" = run_matrix_cached(
    "QC-RFSC_HOLDOUT",
    function() run_qc_rfsc_with_controls(x, control_ids = train_qc_ids, meta_df = meta),
    cache_tag = holdout_tag
  ),
  "TIGER" = qccv_tiger_holdout,
  "SERRF" = qccv_serrf_holdout,
  "WINN auto (QC)" = run_matrix_cached(
    "WINN_AUTO_QC_HOLDOUT",
    function() run_winn_with_controls(x, control_ids = train_qc_ids, meta_df = meta, auto_batch = FALSE, parameters = "auto"),
    cache_tag = holdout_tag
  ),
  "WINN auto-batch (QC)" = run_matrix_cached(
    "WINN_AUTOBATCH_QC_HOLDOUT",
    function() run_winn_with_controls(x, control_ids = train_qc_ids, meta_df = meta, auto_batch = TRUE, parameters = "auto"),
    cache_tag = holdout_tag
  ),
  "WINN default (no QC)" = method_mats[["WINN default (no QC)"]]
)

qccv_summary <- bind_rows(lapply(method_order_all, function(label) {
  vals <- cv_percent(qccv_holdout_mats[[label]], holdout_qc_ids)
  tibble(method = label, value = as.numeric(vals)) |>
    filter(is.finite(value))
})) |>
  group_by(method) |>
  summarise(
    heldout_qc_cv_mean = mean(value, na.rm = TRUE),
    heldout_qc_cv_sd = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(method)

feature_icc_summary <- bind_rows(lapply(method_order_all, function(label) {
  mat <- method_mats[[label]]
  sample_ids <- intersect(non_qc_ids, colnames(mat))
  feature_ids <- intersect(rownames(GT), rownames(mat))
  if (length(sample_ids) < 6L || length(feature_ids) < 2L) {
    return(tibble(method = label, value = numeric(0)))
  }

  tibble(
    method = label,
    value = vapply(feature_ids, function(fid) {
      calc_icc_a1_pair(
        log1p(as.numeric(GT[fid, sample_ids])),
        log1p(as.numeric(mat[fid, sample_ids])),
        min_n = 6L
      )
    }, numeric(1))
  ) |>
    filter(is.finite(value))
})) |>
  group_by(method) |>
  summarise(
    per_feature_icc_mean = mean(value, na.rm = TRUE),
    per_feature_icc_sd = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(method)

sample_corr_summary <- bind_rows(lapply(method_order_all, function(label) {
  mat <- method_mats[[label]]
  sample_ids <- intersect(non_qc_ids, colnames(mat))
  feature_ids <- intersect(rownames(GT), rownames(mat))
  if (length(sample_ids) < 2L || length(feature_ids) < 20L) {
    return(tibble(method = label, value = numeric(0)))
  }

  tibble(
    method = label,
    value = vapply(sample_ids, function(sid) {
      safe_cor(
        log1p(as.numeric(GT[feature_ids, sid])),
        log1p(as.numeric(mat[feature_ids, sid])),
        method = "pearson",
        min_n = 20L
      )
    }, numeric(1))
  ) |>
    filter(is.finite(value))
})) |>
  group_by(method) |>
  summarise(
    per_sample_pearson_mean = mean(value, na.rm = TRUE),
    per_sample_pearson_sd = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(method)

nonqc_segment_meta <- meta |>
  filter(class != "QC", sample_id %in% colnames(method_mats[["Raw"]])) |>
  arrange(run_order)

run_order_gam_summary <- bind_rows(lapply(method_order_all, function(label) {
  compute_metabolite_segment_gam(
    method_mats[[label]],
    nonqc_segment_meta,
    method_label = label,
    transform_fun = log1p,
    min_obs = 6L,
    k_max = 5L
  )
})) |>
  filter(is.finite(explained)) |>
  summarise_metabolite_segment_gam() |>
  rename(
    residual_run_order_gam_mean = mean_explained,
    residual_run_order_gam_sd = sd_explained
  ) |>
  select(method, residual_run_order_gam_mean, residual_run_order_gam_sd) |>
  arrange(method)

ljungbox_summary <- bind_rows(lapply(method_order_all, function(label) {
  compute_metabolite_segment_ljung_box(
    method_mats[[label]],
    nonqc_segment_meta,
    method_label = label,
    transform_fun = log1p,
    min_obs = 8L,
    max_lag = 10L
  )
})) |>
  annotate_metabolite_segment_ljung_box(alpha = 0.05, adjust_method = "BH") |>
  summarise_metabolite_segment_ljung_box() |>
  transmute(
    method,
    autocorrelated_metabolite_plate_count = n_autocorrelated
  ) |>
  arrange(method)

metric_summary <- tibble(method = method_order_all) |>
  left_join(qccv_summary, by = "method") |>
  left_join(feature_icc_summary, by = "method") |>
  left_join(sample_corr_summary, by = "method") |>
  left_join(run_order_gam_summary, by = "method") |>
  left_join(ljungbox_summary, by = "method") |>
  mutate(method = factor(method, levels = method_order_all)) |>
  arrange(method)
```

## 7. Summary

```{r summary}
knitr::kable(
  metric_summary |>
    mutate(across(where(is.numeric), ~ round(.x, 4))),
  caption = "Simulation benchmark summary."
)

holdout_summary <- tibble(
  holdout_seed = 20260311L,
  n_holdout_qc = length(holdout_qc_ids),
  holdout_qc_ids = paste(holdout_qc_ids, collapse = ", ")
)

knitr::kable(holdout_summary, caption = "Held-out QC split used for QC-CV evaluation.")
```
