---
title: "WaveICA 2.0 adenocarcinoma plasma public correction benchmark"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
repo_root <- normalizePath(".", mustWork = TRUE)
if (!file.exists(file.path(repo_root, "scripts", "run_human_public_benchmark.R"))) {
  repo_root <- normalizePath("..", mustWork = TRUE)
}
result_dir <- file.path(repo_root, "results", "waveica_adenocarcinoma")
report_dir <- file.path(repo_root, "reports")
read_result <- function(name) read.csv(file.path(result_dir, name), check.names = FALSE, stringsAsFactors = FALSE)
metrics <- read_result("method_metrics.csv")
metrics_common <- read_result("method_metrics_common_panel.csv")
metrics_coverage <- read_result("method_metrics_coverage_penalized.csv")
dimensions <- read_result("method_dimensions.csv")
runtime <- read_result("method_runtime.csv")
parameters <- read_result("method_parameters.csv")
holdout <- read_result("qc_holdout_ids.csv")
training <- read_result("qc_training_ids.csv")
```

# Scope and provenance

This notebook reports the executed benchmark of the exact `Amide_data.rda`
object from the pinned WaveICA 2.0 source repository. The matrix is the
nonnegative, XCMS-derived input before WaveICA correction. Only this processed
object and small source/documentation files were downloaded; raw spectra were
not downloaded or processed.

```{r audit, results='asis', echo=FALSE}
cat(paste(readLines(file.path(report_dir, "waveica_adenocarcinoma_suitability.md"), warn = FALSE), collapse = "\n"))
```

# Fixed hidden-QC protocol

Fifteen interior pooled QCs were selected with seed 20260811, five per supplied
batch. Each batch retains at least 15 training QCs and its first and last QC.
Held-out identities were replaced by ordinary Sample/Study labels before every
method call and were excluded from all training and parameter selection.

```{r qc-split, echo=FALSE}
knitr::kable(holdout, caption = "Held-out QCs")
knitr::kable(aggregate(sample_id ~ batch, training, length), caption = "Training QCs by batch")
```

# Methods attempted

All nine variants were attempted. Competitor grids and the deterministic
220-feature WiNN tuning panel used training-QC scores only. WiNN auto-batch was
not given supplied batch labels; true batches were retained solely for final
evaluation.

```{r methods, echo=FALSE}
knitr::kable(runtime, digits = 3, caption = "Method status and runtime")
knitr::kable(parameters, caption = "Selected or fixed parameters")
knitr::kable(dimensions, caption = "Output coverage and alignment")
```

```{r tiger-output-audit, echo=FALSE}
tiger_audit_path <- file.path(result_dir, "tiger_nonfinite_output_features.csv")
if (file.exists(tiger_audit_path)) {
  tiger_audit <- read.csv(tiger_audit_path, check.names = FALSE, stringsAsFactors = FALSE)
  if (nrow(tiger_audit)) {
    cat(sprintf("TIGER produced non-finite values in %d complete feature rows. Those rows were removed from TIGER's output, with no cell replacement; all IDs and counts are saved below.\n\n", nrow(tiger_audit)))
    knitr::kable(tiger_audit, caption = "Explicit TIGER non-finite output-row audit")
  }
}
```

# Primary results

Technical entries use every feature retained by each method. Biological
weighted-PC, association, resampling, and cross-batch-effect analyses use the
saved common feature intersection so all method comparisons have identical
feature and patient denominators.

```{r primary-metrics, echo=FALSE}
shown <- c("method", "heldout_qc_cv_mean", "residual_gam_deviance_mean",
           "residual_ljung_box_proportion", "batch_weighted_pc_r2",
           "biological_weighted_pc_r2_mean", "biological_associated_features",
           "cross_batch_effect_pearson_median", "total_runtime_sec",
           "features_retained")
knitr::kable(metrics[, shown], digits = 4, caption = "Primary retained-feature metrics")
```

## Common and coverage-penalized panels

```{r panels, echo=FALSE}
knitr::kable(metrics_common, digits = 4, caption = "Common retained-feature panel")
knitr::kable(metrics_coverage, digits = 4, caption = "Coverage-penalized full panel")
```

# WiNN selectivity and diagnostic equivalence

```{r winn, echo=FALSE}
knitr::kable(read_result("winn_selectivity_summary.csv"), digits = 4,
             caption = "Features and profiles actually modified by each WiNN mode")
knitr::kable(read_result("winn_instrumentation_validation.csv"), digits = 12,
             caption = "Instrumented-stage versus normal-wrapper matrix validation")
```

# Biological preservation

The primary log1p limma model is `feature ~ biological_group`; the sensitivity
model adds supplied batch. Group effects are also estimated independently in
all three batches. The balanced sensitivity repeatedly includes all 71 members
of the smaller group and 71 randomly selected members of the larger group (100
repetitions, seed 20260815). No biological labels enter method tuning.

```{r biology, echo=FALSE}
assoc <- read_result("waveica_group_associations_primary.csv")
assoc_summary <- aggregate(significant ~ method, assoc, sum)
knitr::kable(assoc_summary, caption = "FDR-significant primary group associations")
knitr::kable(read_result("waveica_cross_batch_effect_concordance.csv"), digits = 4,
             caption = "Batch-pair group-effect concordance")
knitr::kable(read_result("waveica_cross_batch_direction_consistency.csv"), digits = 4,
             caption = "Direction consistency of pooled significant features in all three batches")
stability <- read_result("waveica_balanced_resampling_stability.csv")
stability_summary <- aggregate(selected_at_least_half ~ method, stability, sum)
knitr::kable(stability_summary, caption = "Features selected in at least half of balanced resamples")
```

# Figures

```{r figures, echo=FALSE, out.width='96%'}
figure_names <- c(
  "acquisition_design.pdf", "randomization_diagnostics.pdf",
  "raw_vs_winn_pca.pdf", "technical_comparison.pdf",
  "biological_comparison.pdf", "winn_selectivity.pdf",
  "representative_feature_trajectories.pdf", "waveica_batch_effects.pdf"
)
paths <- file.path(result_dir, "figures", figure_names)
paths <- paths[file.exists(paths)]
knitr::include_graphics(paths, rel_path = FALSE)
```

# Reproducibility and limitations

The mapping of numeric group labels is reported neutrally as `group_0` and
`group_1`: source counts align exactly with chronic-enteritis and colorectal-
cancer totals, respectively, but the deposited object lacks per-row disease
names. The 48,820 exact zeros are preserved because the source does not define
them as missing. Patients are distinct across batches, so sample-replicate
correlation and cross-batch sample ICC are deliberately `NA`.

```{r session, echo=FALSE, results='asis'}
cat("```\n")
cat(paste(readLines(file.path(result_dir, "sessionInfo.txt"), warn = FALSE), collapse = "\n"))
cat("\n```\n")
```
